---
title: "2024 Jobs Estimates"
format: html
params:
    est_year: 2024
jupyter: 
  kernelspec:
    name: "ark"
    language: "R"
    display_name: "Ark R Kernel"
---

# Overview

This document is for producing the `{r} params$est_year` Colorado Jobs Estimates from the Colorado State Demography Office ("SDO").

## Changes

The methodology utilized by the SDO in its 2024 estimates program has been significantly revised for multiple reasons, including:

### BLS Data Availability

Changes in BLS (Bureau of Labor Statistics) data availability require a new primary source of certain employment and income data. In particular, industry specific data is no longer available, and because of this the SDO has decided to instead use the Quarterly Census of Employment and incs (QCEW) and Current Employment Statistics (CES) data as the primary source of employment and income data. This simplifies the analysis and allows for more consistent data across industries.

Additionally the availability of income data has changed from what has historically been available from the BLS. Additional sources of income data instead must be relied upon. The new methodology is outlined in the Base Industry section below.

### Employment Sources and Industry Classification

In order to simplify the methodology and allow for consistency with other data sources the SDO has decided to use the North American Industry Classification System (NAICS) as the primary industry classification system. This allows for consistency with other data sources and simplifies the analysis. Additionally, the SDO has decided to use the public versions of the QCEW and CES data, rather than relying on the microdata at this time, to again simplify the process. If significant fidelity is lost in the new methodology, the SDO may consider using the microdata in the future.

## Loading Data and Packages

```{r}
#| label: setup

library(tidyverse)
library(keyring)
library(DBI)
library(RPostgres)
library(duckdb)
library(dbplyr)
library(censusapi)
library(connections)
library(duckplyr)
library(sdoecon)
library(fredr)
library(googledrive)
library(ipumsr)
library(readxl)
library(imputeTS)
library(sdodemog)
library(httr2)
```

Crosswalks are loaded from the SDO Postgres database, using the generic public (non read-write) connection. Connection keys are stored in the `.Renviron` file, which is not included in this repository for security reasons. The Postgres database is accessed using the `DBI` and `RPostgres` packages. Credentials are stored in the `keyring` package, which allows for secure storage of credentials[^1]. The `duckdb` package is used to create a local DuckDB database for processing the data stored as `data/qcew.duckdb`.

[^1]: See https://keyring.r-lib.org/index.html for more information. Public credentials are stored under the service `"pgsdo"`, for the username `"codemog"`

```{r}
#| label: connect_to_db

# create connections list
conn <- list()


# connect to SDO Postgres database
conn$pgsdo <- sdodemog::connect_to_sdo_db()

# Connect to duckdb database of all QCEW data
conn$qcew <- dbConnect(duckdb::duckdb(), "data/qcew.duckdb")

# Optional - use connections for browsing database in compatabile IDE (e.g., RStudio/Positron)
connections::connection_view(conn$pgsdo)
connections::connection_view(conn$qcew)

# create list for functions
fn <- list()

# create list for files
files <- list()

#copy to google drive using drive id for folder 1bbsGG8Hwf8ZGEryFlyM-CTCfEQPfYyki

files$gdrive_folder <- "1bbsGG8Hwf8ZGEryFlyM-CTCfEQPfYyki"

```

```{r}
#| label: load_crosswalks

# create crosswalks list
crosswalks <- list()

# store crosswalk_region
crosswalks$region <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM econ.crosswalk_region")
)

crosswalks$region_tbl <- crosswalks$region |>
  collect()

```

# Total Estimated Jobs

The first step in the process is to estimate the total number of jobs in Colorado for the year `{r params$est_year}`. This is done using the QCEW and CES data, which provides a comprehensive view of employment in the state. Additionally, an estimate of the total number of sole proprietors is derived using the U.S. Census Bureau's Nonemployer Statistics (NES) data, which provides an estimate of the number of sole proprietors in the state.

## QCEW

First QCEW data is downloaded and processed for all years. QCEW data is accessed using the [BLS QCEW CSV Data Slices](https://www.bls.gov/cew/additional-resources/open-data/csv-data-slices.htm) system for accessing CSVs. Data can be downloaded as CSV files using the URL format https://data.bls.gov/cew/data/api/{YEAR}/{QTR}/area/{FIPS5}.csv, where `{YEAR}` is the year of the analysis, starting in 1990 to current, `{QTR}` is `1-4` for a given quarter, or `a` for annual average, and `{FIPS5}` is the 5 digit state plus county FIPS code. A script is provided to download and process the QCEW data for all years, and is available in `r_scripts/process_qcew.R`. The script downloads the data, processes it, and stores it in a DuckDB database `data/qcew.duckdb` for further analysis. Since the process is time consuming, the script is not run in this document, but is available for reference. A Google Drive folder is used for backing up date including both the [raw data](https://drive.google.com/open?id=1xNJTL7PMTY0wDlIUqWh3eE931s62vbh2&usp=drive_fs), and the [processed DuckDB database]().

```{r}
#| label: process_qcew

# source the processing script (uncomment to source)
# source("r_scripts/process_qcew.R")

```

```{r}
#| label: load_qcew

# create list for storing local qcew info
qcew <- list()

# create a table of raw qcew data
qcew$raw <- read_csv_duckdb("data/qcew_lmi.csv", prudence = "lavish") |>
  rename_all(tolower) |>
  select(
    year = periodyear,
    county_fips,
    countyname = areaname,
    own_code = ownership,
    own_title = ownershipdesc,
    ind_code = code,
    ind_title = title,
    estab,
    avg_emp = avgemp,
    tot_wage = totwage,
    top_emp_av = topempav,
    avg_wk_wage = avgwkwage,
    tax_wage = taxwage,
    contrib:realsuppress
  )

# read in naics hierarchy crosswalk from excel file
crosswalks$naics <- map(
  readxl::excel_sheets(
    "data/crosswalks/bls/qcew-naics-hierarchy-crosswalk.xlsx"
  ),
  \(x) {
    read_excel(
      "data/crosswalks/bls/qcew-naics-hierarchy-crosswalk.xlsx",
      col_types = "text",
      sheet = x
    ) |>
      mutate(naics_version = x, .before = 1)
  }
) |>
  bind_rows()

crosswalks$naics_lvls <- crosswalks$naics |>
  select(
    naics_version,
    ends_with("_code")
  ) |>
  rename(naics2_code = sector_code) |>
  pivot_longer(
    cols = -naics_version,
    names_to = "ind_lvl",
    values_to = "ind_code"
  ) |>
  filter(str_detect(ind_lvl, "naics")) |>
  mutate(ind_lvl = str_remove_all(ind_lvl, "(naics)|(_code)")) |>
  distinct()


```

```{r}
#| label: load_naics_yr_crosswalk

# load crosswalk for NAICS codes
crosswalks$naics_years <- read_csv_duckdb(
  "data/crosswalks/xwalk_qcew_naics_year.csv"
) |>
  mutate(naics_version = paste0("v", as.character(naics_year))) |>
  select(-naics_year) |>
  left_join(
    crosswalks$naics_lvls
  )

```

```{r}
#| label: fill_suppressed_qcew
fn$make_qcew_data <- function() {
  qcew$data <- qcew$raw |>
    arrange(county_fips, ind_code, own_code, year) |>
    # replace s; with NA
    mutate(
      across(
        avg_emp:contrib,
        ~ if_else(estab == 0, '0', .x)
      ),
      across(
        avg_emp:contrib,
        ~ na_if(.x, "s;")
      ),
      across(
        avg_emp:contrib,
        as.numeric
      )
    ) |>
    left_join(
      crosswalks$naics_years,
      by = c("year", "ind_code")
    ) |>
    left_join(
      crosswalks$region_tbl |>
        filter(region <= 14) |>
        select(
          county_fips,
          region_code
        )
    ) |>
    mutate(
      ind_lvl = if_else(
        ind_code %in% c("00", "10"),
        "0",
        ind_lvl
      )
    ) |>
    filter(ind_lvl %in% c("0", "1", "2")) |>
    arrange(year, county_fips, ind_code, own_code) |>
    mutate(
      final_suppress = suppress,
      final_avg_emp = avg_emp,
      override_flag = 0,
      override_reason = NA_character_
    ) |>
    collect()
}
qcew$data <- fn$make_qcew_data()
# qcew$data_no_overrides <- fn$make_qcew_data()

```

## Filling suppressed values

We estimate suppressed values through an iterative process. Prior to this process we first have to check that the suppression flag is consistent, as we replaced "s;" values with NA.

```{r}
#verify that all suppressed values have a supression code == 1
# qcew$raw |>
#   filter(avg_emp == "s;" & suppress != 1) |>
#   count()

```

The process for filling suppressed values is as follows: 1. Create a new table in qcew$suppresss to store the suppressed values.
2. Filter the suppression table to only include those rows that are being updated.
3. Create three new columns in the suppressed table - temp_flag, temp_reason, and temp_value. Set the temp_flag to 1, and provide a reason for the override in temp_reason. Set temp_value to the new value for avg_emp.
4. Join the suppressed values back to the main qcew$data table 5. Update the qcew$data table by replacing avg_emp with temp_value, suppress with 0, the override_flag to be the value from temp_flag, and the override_reason to be the value from temp_reason.
6. Remove the temp columns from the main qcew$data table.

A function has been created to do this process, `fn$apply_suppression_overrides`, which takes the suppressed table as one argument, and the override_reason as a second argument, and applies the overrides to the main qcew\$data table. This function is called after each step in the process to apply the overrides.

```{r}
#| label: create_function_apply_overrides

# fn$suppression_override <- function(suppression_df, output_df) {
#   output_df |>
#     left_join(
#       suppression_df |>
#         select(
#           # any_of(c(
#           year,
#           county_fips,
#           ind_code,
#           own_code,
#           temp_flag,
#           temp_reason,
#           temp_value
#           # ))
#         ) |>
#         mutate(final_suppress = 1)
#     ) |>
#     mutate(
#       final_avg_emp = if_else(!is.na(temp_flag), temp_value, final_avg_emp),
#       final_suppress = if_else(!is.na(temp_flag), 0, final_suppress),
#       override_flag = if_else(!is.na(temp_flag), temp_flag, override_flag),
#       override_reason = if_else(!is.na(temp_flag), temp_reason, override_reason)
#     ) |>
#     select(-starts_with("temp_"))
# }

# fn$check_overrides <- function(df) {
#   df |>
#     filter(final_suppress == 1 | override_flag == 1) |>
#     count(override_reason, sort = TRUE)
# }

```

### Industry Code `00` vs `10`

Prior to starting this procedure industry codes `00` and `10` were compared for equivalency and to determine if one could be used all the time over the other, or if they were circumstantially dependent. This was done by comparing the two industries and looking for differences where one was greater than the other.

```{r}
#| label: compare_ind_codes

# qcew$checks$ind_code <- qcew$data |>
#   filter(ind_code %in% c("00", "10")) |>
#   select(year, county_fips, ind_code, own_code, avg_emp = final_avg_emp) |>
#   pivot_wider(
#     names_from = ind_code,
#     values_from = c(avg_emp),
#     names_glue = "{.value}_{ind_code}"
#   ) |>
#   mutate(
#     diff = avg_emp_10 - avg_emp_00,
#     diff_flag = case_when(
#       is.na(avg_emp_10) & !is.na(avg_emp_00) ~ "10 NA",
#       !is.na(avg_emp_10) & is.na(avg_emp_00) ~ "00 NA",
#       is.na(diff) ~ "Both NA",
#       diff > 0 ~ "10 > 00",
#       diff < 0 ~ "00 > 10",
#       diff == 0 ~ "Equal"
#     )
#   )

# qcew$checks$ind_code |>
#   count(diff_flag, sort = TRUE)

```

They were found to be the same however there were a number of circumstances where 10 was NA and 00 was not, or vice versa. The number with 10 NA was much greater so 00 was used but first the missing NAs available in 10 but not 00 were filled into 00. Code `10` was then filtered out of the data for the remainder of the analysis.

```{r}
#| label: fill_ind_code_00

# qcew$suppress$ind_code <- qcew$checks$ind_code |>
#   filter(diff_flag == "00 NA") |>
#   select(year, county_fips, own_code, temp_value = avg_emp_10) |>
#   mutate(
#     ind_code = "00",
#     temp_flag = 1,
#     temp_reason = "Filled from ind_code 10"
#   )

# qcew$data <- fn$suppression_override(qcew$suppress$ind_code, qcew$data)

# qcew$data <- qcew$data |>
#   filter(ind_code != "10")

# fn$check_overrides(qcew$data)
```

### Government Jobs

As the count is identified as 0, we move on to the next step. Here we check to see that total government jobs are not suppressed for any county at each the state, local and federal level, and that the sum of the ownership groups equals the group total. If these are both the case then we can focus only on private ownership (own_code == 50) as we only look at government totals in our analysis.

```{r}
#| label: suppression_gov

# qcew$suppress$gov <- qcew$data |>
#   filter(
#     own_code %in%
#       c("10", "20", "30", "50") &
#       ind_code %in% c("00")
#   ) |>
#   summarize(
#     total_gov_calc = sum(avg_emp, na.rm = TRUE),
#     total_suppressed = sum(suppress, na.rm = TRUE),
#     .by = c(year, county_fips, ind_code)
#   ) |>
#   left_join(
#     qcew$data |>
#       filter(
#         own_code == "00" &
#           ind_code %in% c("00")
#       ) |>
#       select(year, county_fips, ind_code, total_gov_reported = avg_emp),
#     by = join_by(year, county_fips, ind_code)
#   ) |>
#   mutate(
#     temp_value = total_gov_reported - total_gov_calc,
#     temp_flag = 1,
#     temp_reason = "Government Total Suppressed"
#   ) |>
#   filter(total_suppressed > 0) |>
#   arrange(year, county_fips) |>
#   left_join(
#     qcew$data |>
#       select(year, county_fips, ind_code, own_code, suppress) |>
#       filter(
#         own_code %in%
#           c("10", "20", "30", "50") &
#           ind_code %in% c("00") &
#           suppress == 1
#       ),
#     by = join_by(year, county_fips, ind_code),
#     relationship = "one-to-one"
#   ) |>
#   relocate(own_code, .before = ind_code)

```

While some government groups are suppressed there is only one government type suppressed per county, year, industry code (there are both code "00" and "10" that have totals), and there is a difference between the sum of all ownership classes (State, Local, Federal, Private) and the reported total. As such we can fill in the suppressed government values as the difference between the reported total and the sum of the other ownership classes.

```{r}
#| label: override_gov_suppression

# #apply overrides
# qcew$data <- fn$suppression_override(qcew$suppress$gov, qcew$data)

# # verify that sum of state local federal and private equals total for all groupings
# qcew$data |>
#   filter(
#     own_code %in%
#       c("10", "20", "30", "50") &
#       ind_code %in% c("00")
#   ) |>
#   summarize(
#     total_gov_calc = sum(avg_emp, na.rm = TRUE),
#     total_suppressed = sum(suppress, na.rm = TRUE),
#     .by = c(year, county_fips, ind_code)
#   ) |>
#   left_join(
#     qcew$data |>
#       filter(
#         own_code == "00" &
#           ind_code %in% c("00")
#       ) |>
#       select(year, county_fips, ind_code, total_gov_reported = avg_emp)
#   ) |>
#   mutate(
#     diff = total_gov_reported - total_gov_calc
#   ) |>
#   filter(abs(diff) >= 10) |>
#   arrange(year, county_fips)

# # check overrides
# fn$check_overrides(qcew$data)

```

### Private Jobs

A separate data frame is created for the next steps in the process using only private ownership. This is for simplicity in creating and tracking data and will be rejoined with the government data at the end of the process.

```{r}
#| label: filter_private_ownership

# qcew$data_gov <- qcew$data |>
#   filter(own_code != "50")

# fn$calc_industry_gaps <- function(df) {
#   df |>
#     select(
#       -any_of(c(
#         "ind_emp_reported",
#         "ind_emp_calculated",
#         "ind_emp_gap",
#         "ind_supressed"
#       ))
#     ) |>
#     filter(own_code == "50") |>
#     left_join(
#       df |>
#         filter(own_code == "50" & ind_code == "00") |>
#         select(
#           year,
#           county_fips,
#           ind_emp_reported = final_avg_emp
#         )
#     ) |>
#     left_join(
#       df |>
#         filter(own_code == "50" & !ind_code == "00") |>
#         summarize(
#           ind_emp_calculated = sum(final_avg_emp, na.rm = TRUE),
#           ind_supressed = sum(final_suppress, na.rm = TRUE),
#           .by = c(year, county_fips)
#         )
#     ) |>
#     mutate(ind_emp_gap = ind_emp_reported - ind_emp_calculated) |>
#     arrange(county_fips, ind_code, year)
# }

# qcew$data_private <- fn$calc_industry_gaps(qcew$data)

```

#### No job gap

First we check to see if there are any counties where the total private jobs (ind_code == "00") is not suppressed, and the sum of all other industries equals the total. If this is the case then any suppressed values can be filled with 0 jobs. This is done to avoid accidentally filling jobs in later steps when there is no job gap to fill.

```{r}
#| label: suppression_adjust_no_gap

# qcew$suppress$no_gap <- qcew$data_private |>
#   filter(ind_emp_gap <= 0 & ind_supressed > 0 & final_suppress == 1) |>
#   select(year, county_fips, ind_code, own_code) |>
#   mutate(
#     temp_value = 0,
#     temp_flag = 1,
#     temp_reason = "No Job Gap - Filled with 0"
#   )

# #apply overrides
# qcew$data_private <- fn$suppression_override(
#   qcew$suppress$no_gap,
#   qcew$data_private
# )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# fn$check_overrides(qcew$data_private)

```

#### Zero Establishments

Next we replace all suppressed values with 0 establishments to have zero employees, and then if there is only one remaining missing NAICS value for that area we fill that as the difference between the parent total and the sum of the other industries. Columns are added to indicate if suppression was overwritten, the override reason, and the new value. Grouped data frames using group_by are used, rather than within mutate, as the same groupings will be used for the next number of steps.

For now we are only calculating these based on the own_code

```{r}
#| label: suppression_adjust_0_estab

# qcew$suppress$no_estabs <- qcew$data_private |>
#   filter(own_code == "50" & ind_supressed == "1" & final_suppress == "1") |>
#   mutate(
#     temp_value = max(ind_emp_gap, 0),
#     temp_flag = 1,
#     temp_reason = "Only Missing Industry",
#     .by = c(year, county_fips, ind_code)
#   ) |>
#   select(
#     year,
#     county_fips,
#     ind_code,
#     own_code,
#     final_suppress,
#     temp_value,
#     temp_flag,
#     temp_reason
#   )

# #apply overrides
# qcew$data_private <- fn$suppression_override(
#   qcew$suppress$no_estabs,
#   qcew$data_private
# )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # check overrides
# fn$check_overrides(qcew$data_private)

```

### Interpolation

Next we do a simple linear interpolation for any remaining suppressed values, using the imputeTS::na_interpolation function with a max_gap of 2 (so this does not fill overly long gaps).

```{r}
#| label: simple_interpolation

# qcew$suppress$simple_interp <- qcew$data_private |>
#   arrange(county_fips, ind_code, own_code, year) |>
#   mutate(
#     na_count = sum(is.na(final_avg_emp)),
#     n_years = n(),
#     all_na = na_count >= n_years - 1,
#     .by = c(county_fips, ind_code, own_code)
#   ) |>
#   filter(!all_na) |>
#   mutate(
#     init_value = if_else(
#       final_suppress == 1,
#       na_interpolation(final_avg_emp, option = "linear", maxgap = 5),
#       NA
#     ),
#     temp_flag = if_else(!is.na(init_value), 1, NA_real_),
#     temp_reason = if_else(
#       !is.na(init_value),
#       "Simple Linear Interpolation",
#       NA_character_
#     ),
#     .by = c(county_fips, ind_code)
#   ) |>
#   # adjust for county totals
#   mutate(
#     init_total = sum(init_value, na.rm = TRUE),
#     init_share = if_else(
#       init_total > 0,
#       init_value / init_total,
#       0
#     ),
#     init_count = sum(temp_flag == 1, na.rm = TRUE),
#     init_suppress = sum(final_suppress),
#     init_allocate_all = init_count == init_suppress,
#     .by = c(year, county_fips)
#   ) |>
#   #adjust temp_value to not be larger than the employment gap
#   mutate(
#     temp_value = round(
#       if_else(
#         !is.na(init_value) & (init_allocate_all | init_total >= ind_emp_gap),
#         ind_emp_gap * init_share,
#         init_value
#       ),
#       0
#     ),
#   ) |>
#   select(
#     year,
#     county_fips,
#     countyname,
#     own_code,
#     ind_code,
#     ind_title,
#     estab,
#     final_avg_emp,
#     ind_emp_reported,
#     ind_emp_calculated,
#     ind_emp_gap,
#     final_suppress,
#     temp_value,
#     temp_flag,
#     temp_reason,
#     starts_with("init_")
#   ) |>
#   arrange(
#     county_fips,
#     ind_code,
#     year
#   ) |>
#   filter(temp_flag == 1) |>
#   mutate(
#     final_total = sum(temp_value, na.rm = TRUE),
#     .by = c(year, county_fips),
#     .after = ind_emp_gap
#   )

# #apply overrides
# qcew$data_private <- fn$suppression_override(
#   qcew$suppress$simple_interp,
#   qcew$data_private
# )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # fill 1 remaining suppressed gaps
# qcew$data_private <- qcew$data_private |>
#   mutate(
#     final_avg_emp = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       ind_emp_gap,
#       final_avg_emp
#     ),
#     override_flag = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       1,
#       override_flag
#     ),
#     override_reason = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       "Fill Single Industry - Interpolation",
#       override_reason
#     ),
#     final_suppress = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       0,
#       final_suppress
#     ),
#   )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # check overrides
# fn$check_overrides(qcew$data_private)

```

### Region Shares

Next we calculate the share of employment by industry within the region, and allocate the remaining job gaps based on those shares.

```{r}
#| label: region_shares

# # calculate region shares
# qcew$suppress$region_shares <- qcew$data_private |>
#   filter(ind_supressed == 0) |>
#   summarize(
#     ind_emp_region = sum(final_avg_emp, na.rm = FALSE),
#     total_emp_region = sum(ind_emp_reported, na.rm = FALSE),
#     ind_share_region = ind_emp_region / total_emp_region,
#     .by = c(year, ind_code, region_code)
#   ) |>
#   arrange(year, region_code, ind_code) |>
#   mutate(
#     total_industries = n(),
#     .by = c(year, region_code)
#   )

# # check that region shares sum to 1
# qcew$suppress$region_shares |>
#   filter(ind_code != "00") |>
#   summarize(
#     total = sum(ind_share_region, na.rm = TRUE),
#     .by = c(year, region_code)
#   ) |>
#   filter(abs(total - 1) > 0.01)

# # allocate remaining gaps based on region shares
# qcew$suppress$region_allocate <- qcew$data_private |>
#   filter(final_suppress == 1) |>
#   left_join(
#     qcew$suppress$region_shares,
#     by = c("year", "ind_code", "region_code")
#   ) |>
#   mutate(
#     temp_value = round(ind_emp_gap * ind_share_region, 0),
#     temp_flag = if_else(!is.na(temp_value), 1, NA_real_),
#     temp_reason = if_else(
#       !is.na(temp_value),
#       "Region Share Allocation",
#       NA_character_
#     ),
#     .by = c(year, county_fips)
#   ) |>
#   select(
#     year,
#     county_fips,
#     ind_code,
#     own_code,
#     final_suppress,
#     temp_value,
#     temp_flag,
#     temp_reason
#   ) |>
#   filter(!is.na(temp_value))

# #apply overrides
# qcew$data_private <- fn$suppression_override(
#   qcew$suppress$region_allocate,
#   qcew$data_private
# )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # fill 1 remaining suppressed gaps
# qcew$data_private <- qcew$data_private |>
#   mutate(
#     final_avg_emp = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       ind_emp_gap,
#       final_avg_emp
#     ),
#     override_flag = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       1,
#       override_flag
#     ),
#     override_reason = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       "Fill Single Industry - Region Share",
#       override_reason
#     ),
#     final_suppress = if_else(
#       final_suppress == 1 & ind_supressed == 1,
#       0,
#       final_suppress
#     ),
#   )

# # update industry gaps
# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # check overrides
# fn$check_overrides(qcew$data_private)

```

Finally this is repeated with state shares for the remaining gaps.

```{r}
#| label: state_shares

# # calculate state shares
# qcew$suppress$state_shares <- qcew$data_private |>
#   filter(county_fips == "000") |>
#   summarize(
#     ind_emp_state = sum(final_avg_emp, na.rm = FALSE),
#     total_emp_state = sum(ind_emp_reported, na.rm = FALSE),
#     ind_share_state = ind_emp_state / total_emp_state,
#     .by = c(year, ind_code)
#   ) |>
#   arrange(year, ind_code) |>
#   mutate(
#     total_industries = n(),
#     .by = c(year)
#   )

# # check that state shares sum to 1
# qcew$suppress$state_shares |>
#   filter(ind_code != "00") |>
#   summarize(total = sum(ind_share_state, na.rm = TRUE), .by = c(year)) |>
#   filter(abs(total - 1) > 0.01)

# # allocate remaining gaps based on state shares
# qcew$suppress$state_allocate <- qcew$data_private |>
#   filter(final_suppress == 1) |>
#   left_join(
#     qcew$suppress$state_shares,
#     by = c("year", "ind_code")
#   ) |>
#   mutate(
#     temp_value = round(ind_emp_gap * ind_share_state, 0),
#     temp_flag = if_else(!is.na(temp_value), 1, NA_real_),
#     temp_reason = if_else(
#       !is.na(temp_value),
#       "State Share Allocation",
#       NA_character_
#     ),
#     .by = c(year, county_fips)
#   ) |>
#   select(
#     year,
#     county_fips,
#     ind_code,
#     own_code,
#     final_suppress,
#     temp_value,
#     temp_flag,
#     temp_reason
#   ) |>
#   filter(!is.na(temp_value))

# #apply overrides
# qcew$data_private <- fn$suppression_override(
#   qcew$suppress$state_allocate,
#   qcew$data_private
# )

# # update industry gaps

# qcew$data_private <- fn$calc_industry_gaps(qcew$data_private)

# # check overrides

# fn$check_overrides(qcew$data_private)

```

# Sole Proprietors

```{r}
#| label: load_nes_years

# create nes list
nes <- list()

# get available years using censusapi
nes$info <- listCensusApis() |>
  filter(name == "nonemp")

nes$yrs$raw <- nes$info |>
  pull(vintage)

```

Sole proprietor estimates are derived from the U.S. Census Bureau's [Nonemployer Statistics (NES)](https://www.census.gov/programs-surveys/nonemployer-statistics.html) data, which provides an estimate of the number of sole proprietors by county and NAICS. Data is access using the [NES API](https://www.census.gov/data/developers/data-sets/nonemp-api.html). The NES data is released with a lag, and as of the current analysis the most recent available year is `{r} max(nes$yrs$raw)`. The data is retrieved using the [`censusapi`](https://www.hrecht.com/censusapi/) package, which provides a convenient interface to the Census API. The NES data is processed to create a table of sole proprietors by county and NAICS code. The NAICS code is renamed to `ind_code`[^2] to match the QCEW data, and use the `own_code` of `1` to indicate sole proprietors. The NES data is then stored in the DuckDB database for further analysis.

[^2]: As there are not strict concordances between NAICS years we leave these classified as they were at the time - as with the QCEW. At the industry level there should be minimal changes. The one exception to align with the QCEW is that the QCEW utilizes 2002 NAICS codes for all years prior to . 1997 to 2002 concordances were used, allocating evenly.

```{r}
#| label: load_nes

# load NES variables for most recent year
nes$vars <- listCensusMetadata(
  name = "nonemp",
  vintage = max(nes$yrs$raw),
  type = "variables"
)

fn$get_nes_data <- function(vintage) {
  # get list of variables for vintage
  nes$vars <- listCensusMetadata(
    name = "nonemp",
    vintage = vintage,
    type = "variables"
  ) |>
    filter(
      is.na(predicateOnly) &
        !(name %in% c("STATE", "COUNTY"))
    ) |>
    pull(name)

  # get NES data for a given year
  data <- getCensus(
    name = "nonemp",
    vintage = vintage,
    vars = nes$vars,
    region = "county:*",
    regionin = "state:08"
  ) |>
    mutate(
      vintage = vintage,
      .before = 1
    ) |>
    # rename variables to lowercase
    rename_with(tolower) |>
    # rename naics to match QCEW
    rename_with(
      ~"ind_code",
      matches("naics\\d{2,4}$")
    )

  return(data)
}

nes$raw <- tbl(conn$qcew, "nes") |>
  distinct() |>
  arrange(vintage, state, industry_code, county) |>
  collect()

nes$yrs$db <- nes$raw |>
  pull(vintage) |>
  unique()

nes$yrs$qry <- nes$info |>
  select(vintage) |>
  filter(vintage >= 2000 & !(vintage %in% nes$yrs$db)) |>
  pull(vintage) |>
  unique()

# Store NES data in DuckDB
dbWriteTable(
  conn$qcew,
  "nes",
  map(
    nes$yrs$qry,
    fn$get_nes_data,
    .progress = TRUE
  ) |>
    bind_rows(),
  append = TRUE
)

# Create crosswalk for NES data
crosswalks$nes <- nes$raw |>
  collect() |>
  select(industry_code, sector, subsector, indlevel) |>
  distinct() |>
  arrange(industry_code)


# Create table of relevant NES data
nes$data <- nes$raw |>
  mutate(
    area_fips = paste0(state, county),
    own_code = "5", # treat sole props as private,
    industry_code = if_else(
      industry_code == "00",
      "10",
      industry_code
    ),
    year = vintage
  ) |>
  rename(
    ind_code = industry_code
  ) |>
  left_join(
    crosswalks$naics_years
  ) |>
  select(
    year = vintage,
    area_fips,
    own_code,
    ind_code,
    ind_lvl,
    prop = nestab,
    inc = nrcptot,
    f_prop = nestab_f,
    f_inc = nrcptot_f
  ) %>%
  # distinct() |>
  bind_rows(
    . |>
      mutate(
        area_fips = "08000", # Colorado total
      ) |>
      summarize(
        across(
          starts_with(c("prop", "inc")),
          ~ sum(.x, na.rm = FALSE)
        ),
        .by = c("year", "area_fips", "ind_code", "ind_lvl", "own_code")
      )
  ) |>
  arrange(year, area_fips, ind_code, own_code)


```

# Additional Series

Additional series are added to the estimates as they become available. These series include 
- Agriculture estimates from the Census of Ag
- Military Employment estimates from the Department of Defense

# Census of Agriculture
Data from the [Census of Agriculture](https://www.nass.usda.gov/Publications/AgCensus/2022/index.php) is accessed using the USDA NASS API. An API key is required to access the data, which can be obtained from the [NASS Quick Stats API](https://quickstats.nass.usda.gov/api). The API key is stored in the `.Renviron` file for secure access. A function is created to retrieve data from the NASS API based on various parameters such as source description, sector description, group description, commodity description, statistic category description, state alpha code, aggregation level description, and year. The data is retrieved in CSV format and read into a data frame for further analysis.

``` {r}
#| label: load_census_of_ag
source("r_scripts/nass_scripts.R")

```

Data is pulled to get the total number of farm owners, and total farm related employment by county for Colorado for all available Census of Ag years.

```{r}
#| label: get_census_of_ag_data

census_of_ag <- read_csv_duckdb(
  "data/inputs/census_of_ag/*",
  prudence = "lavish"
) |>
  filter(STATE_ALPHA == "CO")


ag_prop <- census_of_ag |>
  filter(
    str_detect(SHORT_DESC, "PRIMARY OCCUPATION, FARMING"),
    str_detect(CLASS_DESC, "ALL"),
    # str_detect(UNIT_DESC, "OPERAT"),
    str_detect(DOMAIN_DESC, "TOTAL"),
    !(YEAR == 2017 & COMMODITY_DESC == "PRODUCERS")
  ) |>
  collect()
ag_prop <- ag_prop |>
  mutate(
    COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE)
  ) |>
  filter(
    !(COUNTY_CODE == "000" &
      YEAR %in% c("2007", "2012") &
      SHORT_DESC !=
        "OPERATORS, PRINCIPAL, PRIMARY OCCUPATION, FARMING - NUMBER OF OPERATORS"),
    !(COUNTY_CODE == "000" &
      YEAR %in% c("2017") &
      SHORT_DESC !=
        "PRODUCERS, PRINCIPAL, PRIMARY OCCUPATION, FARMING - NUMBER OF PRODUCERS"),
    !(COUNTY_CODE == "000" &
      YEAR %in% c("2022") &
      SHORT_DESC !=
        "PRODUCERS, PRIMARY OCCUPATION, FARMING - NUMBER OF PRODUCERS"),
      !is.na(COUNTY_CODE)
  )



ag_ws <- census_of_ag |>
  filter(
    SECTOR_DESC == "ECONOMICS",
    GROUP_DESC == "EXPENSES",
    COMMODITY_DESC == "LABOR",
    STATISTICCAT_DESC == "WORKERS",
    DOMAIN_DESC == "TOTAL",
    UNIT_DESC == "NUMBER",
    str_detect(CLASS_DESC, "GT 150|GE 150")
  ) |>
  collect()
ag_ws <- ag_ws |>
  mutate(COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE))

# interpolate data to fill in missing years using VALUE column

# Prepare agriculture data: clean values and convert to numeric
# First get all county-year combinations from the census years
census_years <- distinct(ag_prop, YEAR)
all_counties <- distinct(ag_prop, COUNTY_CODE, COUNTY_NAME) |>
  arrange(COUNTY_CODE)

ag_combined <- expand_grid(census_years, all_counties) |>
  left_join(
    ag_prop |>
      mutate(
        VALUE = na_if(
          VALUE,
          "(D)"
        ),
        prop = as.numeric(str_remove_all(VALUE, ","))
      ) |>
      select(YEAR, COUNTY_CODE, COUNTY_NAME, COMMODITY_DESC, SHORT_DESC, prop)
  ) |>
  left_join(
    ag_ws |>
      mutate(
        COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE),
        VALUE = na_if(
          VALUE,
          "(D)"
        ),
        ws = as.numeric(str_remove_all(VALUE, ","))
      ) |>
      select(YEAR, COUNTY_CODE, COUNTY_NAME, ws)
  ) |>
  mutate(
    across(c(prop, ws), ~ replace_na(.x, 0)),
    jobs = prop + ws,
  ) |>
  arrange(COUNTY_CODE, YEAR)

# Check for missing values and "(D)" indicators
ag_combined |>
  filter(is.na(jobs)) |>
  distinct(jobs)

# Interpolate missing values for each county and type combination
ag_interp <- expand_grid(
  YEAR = 2001:2024,
  distinct(ag_combined, COUNTY_CODE, COUNTY_NAME)
) |>
  left_join(ag_combined) |>
  arrange(COUNTY_CODE, YEAR) |>
  mutate(
COUNTY_NAME = if_else(
      COUNTY_CODE == "000",
      "Colorado",
      COUNTY_NAME
    ),
    interp_value = imputeTS::na_interpolation(
      jobs,
      option = "linear",
      maxgap = 24
    ),
    ind_code = "11", # NAICS code for Agriculture, Forestry, Fishing and Hunting
    .by = c(COUNTY_CODE)
  ) |>
  fill(interp_value, .direction = "downup")

# Sum workers and proprietors to get total ag employment
ag_total <- ag_interp |>
  select(
    year = YEAR,
    area_fips = COUNTY_CODE,
    COUNTY_NAME,
    ind_code,
    ag_jobs = interp_value
  )

write_csv(ag_total, "data/outputs/agriculture_employment.csv")
```

# Combined Series

Next the QCEW data is joined with the NES data to create a combined table of employment and income data for both inc and salary and sole proprietor jobs, with the former coming from the QCEW and the latter from the NES.

```{r}
#| label: create_combined_series

# create forecast data list
fcst <- list()

qcew_con <- dbConnect(
  duckdb::duckdb(),
  dbdir = "G:/Shared drives/SDO Econ/qcew/microdata/qcew.db",
  read_only = FALSE
)

qcew$data <- tbl(qcew_con, "qcew_views.total_emp") |>
  rename(
    area_fips = cnty,
    ind_code = naics,
    ind_lvl = lvl,
    ws = employment,
    inc = wages
  ) %>%
  collect() %>%
  bind_rows(
    . |>
      summarize(
        area_fips = "000",
        ws = sum(ws, na.rm = TRUE),
        inc = sum(inc, na.rm = TRUE),
        total_firms = sum(total_firms, na.rm = TRUE),
        largest_firm = max(largest_firm, na.rm = TRUE),
        .by = c("year", "ind_code", "ind_lvl", "own_code")
      )
  ) |>
  arrange(year, area_fips, own_code, ind_code)

# create list of wage and salary job changes to use to advance jobs by one year based on pct change
qcew$changes <- qcew$data |>
  filter(
    ind_lvl <= 3
  ) |>
  select(
    year,
    area_fips,
    ind_code,
    own_code,
    ws
  ) |>
  arrange(year, area_fips, ind_code, own_code) |>
  group_by(area_fips, ind_code, own_code) |>
  mutate(
    area_fips = paste0("08", area_fips),
    ind_code = if_else(ind_code == "00", "10", ind_code),
    ws_change = if_else(lag(ws) == 0, 0, (ws - lag(ws)) / lag(ws)),
    ws_change = if_else(is.na(ws_change), 0, ws_change)
  ) |>
  ungroup() |>
  select(-ws)

# advance proprietor jobs by one year for most recent missing year based on pct change in wage and salary jobs if there is an equivalent industry - otherwise use the last year available
nes$data <- nes$data |>
  bind_rows(
    nes$data |>
      filter(year == max(year)) |>
      mutate(
        year = year + 1
      ) |>
      left_join(
        qcew$changes,
        by = c("area_fips", "ind_code", "own_code", "year")
      ) |>
      mutate(
        prop = prop * (1 + ws_change),
        inc = inc * (1 + ws_change)
      ) |>
      select(-ws_change)
  )

# create combined series of QCEW and NES data for counties
# fcst$tej$
series_c <- qcew$data |>
  relocate(ind_lvl, .after = ind_code) |>
  # add on ag data
  left_join(
    expand_grid(own_code = c("0", "5"), ag_total),
    by = c("year", "area_fips", "ind_code", "own_code")
  ) |>
  mutate(
    ag_jobs = if_else(
      ind_code == "11" & !is.na(ag_jobs),
      ag_jobs,
      0
    )
  ) |>
  mutate(
    area_fips = paste0("08", area_fips),
    own_code = str_sub(own_code, 1, 1),
    ind_code = if_else(ind_code == "00", "10", ind_code),
    ind_lvl = if_else(ind_code == "10", "0", as.character(ind_lvl))
  ) |>
  rename(firmws = total_firms) |>
  filter(
    own_code %in%
      c("1", "2", "3", "5") &
      year >= 2000 &
      ind_lvl %in% c("0", "2", "3")
  ) |>
  full_join(
    nes$data |>
      filter((prop + inc) > 0) |>
      mutate(
        ind_lvl = if_else(ind_code == "10", "0", ind_lvl)
      ),
    by = c("year", "area_fips", "ind_code", "own_code", "ind_lvl"),
    suffix = c("ws", "p")
  ) |>
  mutate(
    area = if_else(
      area_fips == "08000",
      "co",
      paste0("c", str_sub(area_fips, -3))
    ),
    area_lvl = if_else(
      area_fips == "08000",
      "state",
      "county"
    ),
    across(
      starts_with(c("ws", "inc", "prop", "firmws", "largest_firm")),
      \(x) replace_na(x, 0)
    ),
    jobs = ws + prop + ag_jobs,
    inc = incws + incp,
    firms = firmws + prop
  ) |>
  # select(-ag_jobs) |>
  relocate(area, .after = area_fips) %>%
  # add consolidate 0 ownership code series
  bind_rows(
    . |>
      # filter((agglvl_code%%10) %in% c(0, 3:5)) |>
      mutate(
        own_code = "0"
      ) |>
      summarize(
        across(
          starts_with(c("jobs", "ws", "inc", "prop", "firmws", "firms")),
          ~ sum(.x, na.rm = FALSE)
        ),
        across(
          starts_with("largest_firm"),
          ~ max(.x, na.rm = TRUE)
        ),
        .by = c(
          "year",
          "area_fips",
          "area",
          "area_lvl",
          "ind_code",
          "own_code",
          "ind_lvl"
        )
      )
  ) |>
  relocate(starts_with("f_"), .after = last_col()) |>
  arrange(year, area_fips, ind_code, own_code)

# create combined series for Regions
fcst$tej$series_r <- fcst$tej$series_c |>
  mutate(county_fips = str_sub(area_fips, -3)) |>
  inner_join(
    crosswalks$region_tbl |>
      filter(region <= 14) |>
      select(
        county_fips,
        region_code
      )
  ) |>
  mutate(
    area_fips = paste0("08", region_code),
    area_lvl = "region",
    area = paste0("r", region_code)
  ) |>
  summarize(
    across(
      starts_with(c("jobs", "ws", "inc", "prop", "firm")),
      ~ sum(.x, na.rm = TRUE)
    ),
    across(
      starts_with("largest_firm"),
      ~ max(.x, na.rm = TRUE)
    ),
    .by = c(
      "year",
      "area_fips",
      "area",
      "area_lvl",
      "ind_code",
      "own_code",
      "ind_lvl"
    )
  ) |>
  arrange(year, area_fips, ind_code, own_code)

# create combined series for Colorado
fcst$tej$series <- bind_rows(
  fcst$tej$series_c,
  fcst$tej$series_r
)

dbWriteTable(
  conn$qcew,
  "total_estimated_jobs",
  fcst$tej$series,
  overwrite = TRUE
)

dbDisconnect(qcew_con)
dbDisconnect(conn$qcew)

```

# Forecast Preparation

## Total Estimated Jobs

Data next has to be prepared for forecasting in Eviews. This is done at the 2 digit NAICS level, and so data must be output at that level. Government jobs are aggregated into a single industry for each state, local, and federal government, and all other industries are private jobs.

```{r}
#| label: prepare_total_jobs

# load jobs crosswalk from csv
crosswalks$jobs <- sdodemog::pull_db_table("xwalk", "forecast_jobs")

fcst$tej$output <- fcst$tej$series |>
  select(
    -starts_with("f_")
  ) |>
  pivot_longer(
    starts_with(c("jobs", "ws", "inc", "prop")),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    naics = if_else(
      own_code %in% c("1", "2", "3", "8") & ind_lvl == "4",
      "100",
      ind_code
    ),
    .after = ind_code
  ) |>
  inner_join(
    crosswalks$jobs |>
      select(
        naics,
        own_code,
        ind,
        ind_sort
      ) |>
      filter(
        !is.na(ind)
      )
  ) %>%
  bind_rows(
    . |>
      filter(ind %in% c("traderet", "tradewh", "transware", "util")) |>
      mutate(ind = "ttu", naics = "200")
  ) |>
  mutate(
    var = paste(
      variable,
      ind,
      area,
      sep = "_"
    )
  ) |>
  summarize(
    value = sum(value, na.rm = TRUE),
    .by = c(year, area_fips, area_lvl, var, naics) #, naics, own_code, ind_lvl, var)
  ) |>
  arrange(year, area_fips, naics) |>
  pivot_wider(
    names_from = "year",
    values_from = "value"
  )


# Save the outputx to a CSV file
fn$save_outputs <- function(df, lvl) {
  file_loc <- paste0("data/outputs/", lvl, "_jobs_est.csv")

  df |>
    filter(
      area_lvl == lvl
    ) |>
    select(var, matches("\\d{4}$")) |>
    write_csv(file_loc)

  drive_put(
    file_loc,
    path = as_id(files$gdrive_folder),
    name = basename(file_loc),
    type = "csv"
  )
}

map(c("county", "state", "region"), \(x) fn$save_outputs(fcst$tej$output, x))


```

## Moodys Job Estimates

Moodys Analytics provides national forecasts of various economic variables that are used to guide the Colorado forecasts. While most of these are industry job estimates, other variables are also used that may be useful in forecasting jobs at the local level. These variables are downloaded from the [Moodys Analytics Data Buffet](https://www.economy.com/data) and stored in the DuckDB database for further analysis. The data is downloaded using the `get_moodys_series` function, which retrieves the data for a given year and variable. A crosswalk file is used for mapping the Moodys analytics industry codes to the NAICS codes used in the Colorado estimates.

```{r}
# | label: load_moodys_jobs

# get list of moody's employment variables
get_moodys_token()
check_moodys_token()

# pull list of moody's variables from the crosswalk to pipe into pull
fcst$vars$jobs <- crosswalks$jobs |>
  pull(mnemonic)


# add moodys data using sdoecon get_moodys_series function
fcst$moodysjobs$raw <- map(
  #break fcst into chunks of max 25
  expand_grid(
    split(fcst$vars$jobs, seq_along(fcst$vars$jobs) %/% 25),
    as.character(c(1:12))
  ),
  \(x) {
    get_moodys_series(x, vintage = "202508", freq = "204") |>
      convert_moodys()
  }
) |>
  bind_rows()

map(as.character(c(1:12)), \(x) {
  get_moodys_series(
    "FET.IUSA",
    vintage = paste0("2024", x),
    freq = "204"
  ) |>
    convert_moodys() |>
    mutate(
      vintage = paste0("2024", "-", x)
    )
})


sdo_db_write_table(
  fcst$moodysjobs$raw,
  schema = "forecast",
  table = "moodys",
  vintage = "2025-",
  version = "v1",
  overwrite = TRUE,
  latest = TRUE
)

fcst$moodysjobs$data <- pull_db_table(schema = "forecast", table = "moodys") |>
  inner_join(
    crosswalks$jobs |>
      filter(
        !is.na(ind)
      ) |>
      mutate(
        var = paste0("jobs_", ind, "_us")
      ) |>
      select(
        mnemonic,
        var
      ),
    by = c("mnemonic")
  ) |>
  relocate(var, .after = mnemonic) |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

files$moodysjobs <- "data/outputs/moodysjob_vars.csv"

# Save the output to a CSV file
fcst$moodysjobs$data |>
  write_csv(files$moodysjobs)


drive_put(
  files$moodysjobs,
  path = as_id(files$gdrive_folder),
  name = basename(files$moodysjobs),
  type = "csv"
)
```

```{r}
# | label: load_moodys_other

crosswalks$moodys <- pull_db_table("xwalk", "forecast_moodys")

# pull list of moody's variables from the crosswalk to pipe into pull
fcst$vars$moodys <- crosswalks$moodys |>
  pull(mnemonic)


# add moodys data using sdoecon get_moodys_series function
fcst$moodys$raw <- map(
  #break fcst into chunks of max 25
  split(fcst$vars$moodys, seq_along(fcst$vars$moodys) %/% 25),
  \(x) {
    get_moodys_series(x, freq = "204") |>
      convert_moodys()
  }
) |>
  bind_rows()

fcst$moodys$combined <- bind_rows(
  fcst$moodysjobs$raw,
  fcst$moodys$raw
)

sdo_db_write_table(
  fcst$moodys$combined,
  schema = "forecast",
  table = "moodys",
  vintage = "2025-09",
  version = "v1",
  overwrite = TRUE,
  latest = TRUE
)

fcst$moodys$data <- fcst$moodys$raw |>
  inner_join(
    crosswalks$moodys |>
      select(
        mnemonic,
        var
      ),
    by = c("mnemonic")
  ) |>
  relocate(var, .after = mnemonic) |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

files$moodys <- "data/outputs/moodys_vars.csv"

# Save the output to a CSV file
fcst$moodys$data |>
  write_csv(files$moodys)


drive_put(
  files$moodys,
  path = as_id(files$gdrive_folder),
  name = basename(files$moodys),
  type = "csv"
)
```

## FRED Variables

A number of additional economic (non direct jobs) variables are used in the forecasting process, and these are downloaded from the [FRED API](https://fred.stlouisfed.org/). These variables are used to provide additional context for the forecasting process, and are stored in the DuckDB database for further analysis. The `get_fred_series` function from the fredr package is used to retrieve the data for a given year and variable. A crosswalk file is used for mapping FRED variables is stored in "data/crosswalks/xwalk_fred.csv".

```{r}
# | label: load_fred_data

# load fred crosswalk from csv
crosswalks$fred <- read_csv(
  "data/crosswalks/xwalk_fred.csv"
)

# create laus crosswalk
crosswalks$fred_laus <- expand_grid(
  county_fips = unique(crosswalks$region_tbl$county_fips),
  fred_id_num = as.character(3:5)
) |>
  left_join(
    tibble(
      fred_id_num = as.character(3:5),
      var = c("ur", "unem", "empl")
    )
  ) |>
  mutate(
    fred_series = paste0(
      "LAUCN08",
      county_fips,
      "000000000",
      fred_id_num,
      "A"
    )
  ) |>
  arrange(county_fips, fred_id_num)

# pull variables from crosswalk to pipe into pull
fcst$fred$raw <- crosswalks$fred |>
  pmap(
    \(fred_series, var) {
      fredr(
        fred_series,
        frequency = "a",
        aggregation_method = "avg"
      ) |>
        mutate(
          var = var,
          .before = 1
        )
    },
    .progress = TRUE
  ) |>
  bind_rows()
fcst <- list()

# download county level laus data
# fcst$fred$raw_laus <- crosswalks$fred_laus |>
#   filter(
#     county_fips != "014" # exclude Broomfield as it does not have its own series
#   ) |>
#   select(fred_series, var, county_fips) |>
#   pmap(
#     slowly(
#       \(fred_series, var, county_fips) {
#         fredr(
#           fred_series
#         ) |>
#           mutate(
#             county_fips = county_fips,
#             var = var,
#             .before = 1
#           )
#       },
#       rate = rate_delay(0.15)
#     ),
#     .progress = TRUE
#   ) |>
#   bind_rows()

# pull county level laus data from postgres to avoid repeated downloads

fcst$fred$data <- fcst$fred$raw |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

# fcst$fred$data_laus <- fcst$fred$raw_laus |>
#   mutate(
#     year = year(date),
#     #modify var to extract text from between the two "_"
#     var = str_extract(var, "(?<=cps_).*(?=_c)")
#   ) |>
#   select(year, county_fips, var, value)

fcst$fred$county_laus <- sdodemog::pull_db_table(
  schema = "econ",
  table = "laus_county"
)

fcst$fred$region_laus <- sdodemog::pull_db_table(
  schema = "econ",
  table = "laus_region"
)


files$fred <- "data/outputs/fred_vars.csv"

# Save the output to a CSV file
fcst$fred$data |>
  write_csv(files$fred)


drive_put(
  files$fred,
  path = as_id(files$gdrive_folder),
  name = basename(files$fred),
  type = "csv"
)
```

## SDO Variables

```{r}
#| label: load_sdo_vars

# get sdo tables from postgres
sdo <- list()

# load pop by sya
sdo$pop_sya <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.county_sya")
) |>
  collect()

sdo$lfpr <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.labor_force_participation")
) |>
  collect()

sdo$profiles <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.county_profiles")
) |>
  collect()

sdo$household_projections <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.household_projections")
) |>
  collect()

sdo$components_change <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.components_change")
) |>
  collect()


sdo$output <- sdo$pop_sya |>
  filter(countyfips == 0) |>
  mutate(
    age_group = cut(age, c(0, 4, 19, 24, 44, 64, 74, 84, Inf), right = FALSE)
  ) |>
  summarize(
    demo_pop_co = sum(totalpopulation),
    .by = c(year, countyfips, age_group)
  ) |>
  mutate(
    age_group = paste0(
      "pop",
      str_replace_all(as.character(age_group), "\\[|\\)|,|Inf", "")
    )
  ) |>
  pivot_wider(
    names_from = age_group,
    values_from = demo_pop_co
  ) |>
  arrange(year) |>
  left_join(
    sdo$profiles |>
      select(
        countyfips,
        year,
        births,
        deaths,
        netmig = netmigration,
        hh = households,
        hu = totalhousingunits,
        poptot = totalpopulation
      ) |>
      bind_rows(
        sdo$components_change |>
          filter(countyfips == 0 & datatype == "Projection") |>
          select(
            countyfips,
            year,
            births,
            deaths,
            netmig,
            poptot = estimate
          ) |>
          left_join(
            sdo$household_projections |>
              filter(age_group_id == 0 & household_type_id == 0) |>
              select(
                countyfips = area_code,
                year,
                hh = total_households
              ),
            by = c("countyfips", "year")
          )
      )
  ) |>
  left_join(
    sdo$lfpr |>
      rename(year = population_year) |>
      summarize(
        lf = sum(laborforce),
        cni16 = sum(cni_pop_16pl),
        lfpr = lf / cni16,
        .by = c(year)
      )
  ) |>
  arrange(countyfips, year) |>
  pivot_longer(
    cols = -c(year, countyfips),
    names_to = "var",
    values_to = "value"
  ) |>
  mutate(var = paste("demo", var, "co", sep = "_")) |>
  select(-countyfips) |>
  pivot_wider(
    names_from = var,
    values_from = value
  ) |>
  arrange(year)

files$sdo <- "data/outputs/sdo_vars.csv"

sdo$output |>
  write_csv(files$sdo)

drive_put(
  files$sdo,
  path = as_id(files$gdrive_folder),
  name = basename(files$sdo),
  type = "csv"
)

```

## Labor Force Participation Rates by Age

The BLS provides both estimated and projected labor force participation rates by age and sex at the national level, however state and local level data is not available at either the state or local level. In order to estimate labor force participation rates by age and sex for Colorado and counties within it the SDO begins with ACS estimates of labor force participation rates and then applies the BLS projected labor force participation to project forward LFPR.

## LEHD Data

Data from the [LEHD](https://www.census.gov/programs-surveys/lehd.html) program is used to estimate the number of commuters and multiple job holding rates for Colorado. The LEHD data is produced using a separate script, that can be found either on the [Google Drive](https://drive.google.com/drive/folders/1z3nXScGg1KMfK-3bPrmDbOMhmCGlDvmY?usp=drive_link), or on the [process_lehd GitHub repository](https://github.com/gtottenSDO/process_lehd). The Google Drive link is referenced as it already has the necessary files downloaded, which can take some time if not already saved.

Once the LEHD data is downloaded and processed, it is stored both in a duckdb database (found on the Google Drive) and with the final tables saved in Postgres. This program just pulls the final tables from the Postgres database, which can then be loaded into the forecast.

```{r}
#| label: load_lehd

# load LEHD data from postgres into lehd list
lehd <- list()

lehd$commute <- dbReadTable(
  conn$pgsdo,
  Id(
    schema = "econ",
    table = "lehd_net"
  )
)

# load multiple job holding
lehd$mjh <- dbReadTable(
  conn$pgsdo,
  Id(
    schema = "econ",
    table = "lehd_mjh"
  )
)

# combine commute and multiple job holding into single output table
lehd$output <- lehd$commute |>
  summarize(
    county_fips = "co",
    commute = sum(net_commute, na.rm = TRUE),
    .by = c("year")
  ) |>
  left_join(
    lehd$mjh |>
      summarize(
        county_fips = "co",
        primary_jobs = sum(primary_jobs, na.rm = TRUE),
        all_jobs = sum(all_jobs, na.rm = TRUE),
        mjhr = all_jobs / primary_jobs,
        .by = c("year", state_fips)
      ),
    by = c("year", "county_fips")
  ) |>
  filter(state_fips == "08") |>
  arrange(year) |>
  select(
    year,
    county_fips,
    commute,
    mjhr
  ) |>
  pivot_longer(
    cols = -c(year, county_fips)
  ) |>
  mutate(
    var = paste("lehd", name, county_fips, sep = "_")
  ) |>
  select(-county_fips, -name) |>
  pivot_wider(
    names_from = var,
    values_from = value
  )

files$lehd <- "data/outputs/lehd_vars.csv"

lehd$output |>
  write_csv(files$lehd)

drive_put(
  files$lehd,
  path = as_id(files$gdrive_folder),
  name = basename(files$lehd),
  type = "csv"
)
```