---
title: "2024 Jobs Estimates"
format: html
params:
    est_year: 2024
execute:
    eval: false
---

# Overview

This document is for producing the `{r} params$est_year` Colorado Jobs Estimates from the Colorado State Demography Office ("SDO").

## Changes

The methodology utilized by the SDO in its 2024 estimates program has been significantly revised for multiple reasons, including:

### BLS Data Availability

Changes in BLS (Bureau of Labor Statistics) data availability require a new primary source of certain employment and income data. In particular, industry specific data is no longer available, and because of this the SDO has decided to instead use the Quarterly Census of Employment and incs (QCEW) and Current Employment Statistics (CES) data as the primary source of employment and income data. This simplifies the analysis and allows for more consistent data across industries.

Additionally the availability of income data has changed from what has historically been available from the BLS. Additional sources of income data instead must be relied upon. The new methodology is outlined in the Base Industry section below.

### Employment Sources and Industry Classification

In order to simplify the methodology and allow for consistency with other data sources the SDO has decided to use the North American Industry Classification System (NAICS) as the primary industry classification system. This allows for consistency with other data sources and simplifies the analysis. Additionally, the SDO has decided to use the public versions of the QCEW and CES data, rather than relying on the microdata at this time, to again simplify the process. If significant fidelity is lost in the new methodology, the SDO may consider using the microdata in the future.

## Loading Data and Packages

```{r}
#| label: setup

library(tidyverse)
library(keyring)
library(DBI)
library(RPostgres)
library(duckdb)
library(dbplyr)
library(censusapi)
library(connections)
library(duckplyr)
library(sdoecon)
library(fredr)
library(googledrive)
library(ipumsr)
library(readxl)
library(imputeTS)
library(sdodemog)
library(httr2)

```

Crosswalks are loaded from the SDO Postgres database, using the generic public (non read-write) connection. The Postgres database connection is managed using the SDO's internal `sdodemog` package. Credentials are stored in the `keyring` package, which allows for secure storage of credentials[^1]. Crosswalks are stored in a list `xwalk` for reference throughout the process. The `pull_db_table` function from the `sdodemog` package is used to pull crosswalks from the database, and these are stored in the `xwalk` list for reference throughout the process.

The `duckdb` package is used to create a local DuckDB connection to the SDO internal QCEW database found in `G:/Shared drives/SDO File Storage/data_resources/qcew/microdata/qcew.db`. This database contains the QCEW microdata that is provided to the SDO by Colorado Department of Labor and Employment's Labor Market Information (LMI) unit. The data is accessed using SQL queries, and processed using the `duckplyr` package for efficient processing of large datasets. Prior to running these scripts the QCEW database needs to be updated. This is done by running the necessary SQL scripts located in the `G:/Shared drives/SDO File Storage/data_resources/qcew/microdata/` folder. Detailed instructions can be found in that documentation. The `connections` package is used to manage database connections and allow for browsing of the databases in compatible IDEs such as RStudio/Positron.

[^1]: See https://keyring.r-lib.org/index.html for more information. Public credentials are stored under the service `"pgsdo"`, for the username `"codemog"`. Public credentials are read only, and restricted to what databases they can access. Users looking to write data should utilize their own internal individual credentials, which should include the necessary read/write privilages, as provided by the SDO staff.

```{r}
#| label: connect_to_db

# create connections list
conn <- list()


# connect to SDO Postgres database
conn$pgsdo <- sdodemog::connect_to_sdo_db()

# Connect to duckdb database of all QCEW data
conn$qcew <- dbConnect(
  duckdb::duckdb(),
  # storage location for duckdb database of all QCEW data - internal - change directory as needed - recommenced to save file down prior to running
  dbdir = if (interactive()) file.choose() else "G:/Shared drives/SDO File Storage/data_resources/qcew/microdata/qcew.db",
  read_only = TRUE
)

# Optional - use connections for browsing database in compatabile IDE (e.g., RStudio/Positron)
connections::connection_view(conn$pgsdo)
connections::connection_view(conn$qcew)

# create list for functions
fn <- list()

# create list for files
files <- list()

#copy to google drive using drive id for folder 1bbsGG8Hwf8ZGEryFlyM-CTCfEQPfYyki

files$gdrive_folder <- "1bbsGG8Hwf8ZGEryFlyM-CTCfEQPfYyki"

```

```{r}
#| label: load_crosswalks

# store crosswalk_region
xwalk_region <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM econ.crosswalk_region")
)

xwalk_region_tbl <- xwalk_region |>
  collect()


# read in naics hierarchy crosswalk from excel file
xwalk_naics <- pull_db_table("xwalk", "naics")

# load crosswalk for NAICS codes
xwalk_naics_years <- pull_db_table("xwalk", "naics_long_year")


```

# Total Estimated Jobs

The first step in the process is to estimate the total number of jobs in Colorado for the year `{r params$est_year}`. This is done using the QCEW and CES data, which provides a comprehensive view of employment in the state. Additionally, an estimate of the total number of sole proprietors is derived using the U.S. Census Bureau's Nonemployer Statistics (NES) data, which provides an estimate of the number of sole proprietors in the state.

## QCEW

QCEW data is comes from the QCEW microdata provided to the SDO by LMI. Full documentation can be found for the QCEW processing, but the general methodology is to assign a full NAICS crosswalk to the 6 digit industry NAICS provided by LMI, and calculate the total annual jobs by county and NAICS code for all private industries. Government jobs are also aggregated, and assigned their own NAICS codes for analysis, breaking them into local government, state government, and federal government. The QCEW data is processed using SQL queries in the DuckDB database, and then further processed in R to create a combined series of employment and income data for both wage and salary jobs and sole proprietor jobs.

The processed data included in this database has not yet been supressed and so *SHOULD NOT BE SHARED OUTSIDE OF THE SDO* without further processing to suppress data as needed. The data stored in the table `qcew_views.total_emp` is the fully processed data, and includes data on industry total employment (`ws`) and wages (`inc`), as well as total firms, and largest firm size. Employment is used to calculate total employment and wages are included for calculations in assigning jobs to base industries. 

``` {r}
#| label: process_qcew

qcew_data <- tbl(conn$qcew, "qcew_views.estimates_emp") |> 
  rename(ws = employment)

# Check to make sure there are not duplicate records for any year, area_id, and naics combinations
qcew_data |> 
  count(year, area_id, naics) |> 
  filter(n > 1)

# create list of wage and salary job changes to use to advance jobs by one year based on pct change
qcew_changes <- qcew_data |>
  filter(
    lvl <= 3
  ) |>
  select(
    year,
    area_id,
    naics,
    ws
  ) |>
  arrange(year, area_id, naics) |>
  group_by(area_id, naics) |>
  mutate(
    area_id = paste0("08", area_id),
    naics = if_else(naics == "00", "10", naics),
    ws_change = if_else(lag(ws) == 0, 0, (ws - lag(ws)) / lag(ws)),
    ws_change = if_else(is.na(ws_change), 0, ws_change)
  ) |>
  ungroup() |>
  select(-ws)

```

# Sole Proprietors


Sole proprietor estimates are derived from the U.S. Census Bureau's [Nonemployer Statistics (NES)](https://www.census.gov/programs-surveys/nonemployer-statistics.html) data, which provides an estimate of the number of sole proprietors by county and NAICS. Data is access using the [NES API](https://www.census.gov/data/developers/data-sets/nonemp-api.html). The NES data is released with a lag, and as of the current analysis the most recent available year is `{r} max(nes$yrs$raw)`. The data is retrieved using the [`censusapi`](https://www.hrecht.com/censusapi/) package, which provides a convenient interface to the Census API. The NES data is processed to create a table of sole proprietors by county and NAICS code. The NAICS code is renamed to `naics`[^2] to match the QCEW data, and use the `own_code` of `1` to indicate sole proprietors. The NES data is then stored in the DuckDB database for further analysis.

[^2]: As there are not strict concordances between NAICS years we leave these classified as they were at the time - as with the QCEW. At the industry level there should be minimal changes. The one exception to align with the QCEW is that the QCEW utilizes 2002 NAICS codes for all years prior to . 1997 to 2002 concordances were used, allocating evenly.

```{r}
#| label: load_nes



```

# Additional Series

Additional series are added to the estimates as they become available. These series include - Agriculture estimates from the Census of Ag - Military Employment estimates from the Department of Defense

# Census of Agriculture

Data from the [Census of Agriculture](https://www.nass.usda.gov/Publications/AgCensus/2022/index.php) is accessed using the USDA NASS API. An API key is required to access the data, which can be obtained from the [NASS Quick Stats API](https://quickstats.nass.usda.gov/api). The API key is stored in the `.Renviron` file for secure access. A function is created to retrieve data from the NASS API based on various parameters such as source description, sector description, group description, commodity description, statistic category description, state alpha code, aggregation level description, and year. The data is retrieved in CSV format and read into a data frame for further analysis.

```{r}
#| label: load_census_of_ag
# source("r_scripts/nass_scripts.R")

```

Data is pulled to get the total number of farm owners, and total farm related employment by county for Colorado for all available Census of Ag years.

```{r}
#| label: get_census_of_ag_data

# census_of_ag <- read_csv_duckdb(
#   "data/inputs/census_of_ag/*",
#   prudence = "lavish"
# ) |>
#   filter(STATE_ALPHA == "CO")


# ag_prop <- census_of_ag |>
#   filter(
#     str_detect(SHORT_DESC, "PRIMARY OCCUPATION, FARMING"),
#     str_detect(CLASS_DESC, "ALL"),
#     # str_detect(UNIT_DESC, "OPERAT"),
#     str_detect(DOMAIN_DESC, "TOTAL"),
#     !(YEAR == 2017 & COMMODITY_DESC == "PRODUCERS")
#   ) |>
#   collect()
# ag_prop <- ag_prop |>
#   mutate(
#     COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE)
#   ) |>
#   filter(
#     !(COUNTY_CODE == "000" &
#       YEAR %in% c("2007", "2012") &
#       SHORT_DESC !=
#         "OPERATORS, PRINCIPAL, PRIMARY OCCUPATION, FARMING - NUMBER OF OPERATORS"),
#     !(COUNTY_CODE == "000" &
#       YEAR %in% c("2017") &
#       SHORT_DESC !=
#         "PRODUCERS, PRINCIPAL, PRIMARY OCCUPATION, FARMING - NUMBER OF PRODUCERS"),
#     !(COUNTY_CODE == "000" &
#       YEAR %in% c("2022") &
#       SHORT_DESC !=
#         "PRODUCERS, PRIMARY OCCUPATION, FARMING - NUMBER OF PRODUCERS"),
#     !is.na(COUNTY_CODE)
#   )


# ag_ws <- census_of_ag |>
#   filter(
#     SECTOR_DESC == "ECONOMICS",
#     GROUP_DESC == "EXPENSES",
#     COMMODITY_DESC == "LABOR",
#     STATISTICCAT_DESC == "WORKERS",
#     DOMAIN_DESC == "TOTAL",
#     UNIT_DESC == "NUMBER",
#     str_detect(CLASS_DESC, "GT 150|GE 150")
#   ) |>
#   collect()
# ag_ws <- ag_ws |>
#   mutate(COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE))

# # interpolate data to fill in missing years using VALUE column

# # Prepare agriculture data: clean values and convert to numeric
# # First get all county-year combinations from the census years
# census_years <- distinct(ag_prop, YEAR)
# all_counties <- distinct(ag_prop, COUNTY_CODE, COUNTY_NAME) |>
#   arrange(COUNTY_CODE)

# ag_combined <- expand_grid(census_years, all_counties) |>
#   left_join(
#     ag_prop |>
#       mutate(
#         VALUE = na_if(
#           VALUE,
#           "(D)"
#         ),
#         prop = as.numeric(str_remove_all(VALUE, ","))
#       ) |>
#       select(YEAR, COUNTY_CODE, COUNTY_NAME, COMMODITY_DESC, SHORT_DESC, prop)
#   ) |>
#   left_join(
#     ag_ws |>
#       mutate(
#         COUNTY_CODE = if_else(AGG_LEVEL_DESC == "STATE", "000", COUNTY_CODE),
#         VALUE = na_if(
#           VALUE,
#           "(D)"
#         ),
#         ws = as.numeric(str_remove_all(VALUE, ","))
#       ) |>
#       select(YEAR, COUNTY_CODE, COUNTY_NAME, ws)
#   ) |>
#   mutate(
#     across(c(prop, ws), ~ replace_na(.x, 0)),
#     jobs = prop + ws,
#   ) |>
#   arrange(COUNTY_CODE, YEAR)

# # Check for missing values and "(D)" indicators
# ag_combined |>
#   filter(is.na(jobs)) |>
#   distinct(jobs)

# # Interpolate missing values for each county and type combination
# ag_interp <- expand_grid(
#   YEAR = 2001:2024,
#   distinct(ag_combined, COUNTY_CODE, COUNTY_NAME)
# ) |>
#   left_join(ag_combined) |>
#   arrange(COUNTY_CODE, YEAR) |>
#   mutate(
#     COUNTY_NAME = if_else(
#       COUNTY_CODE == "000",
#       "Colorado",
#       COUNTY_NAME
#     ),
#     interp_value = imputeTS::na_interpolation(
#       jobs,
#       option = "linear",
#       maxgap = 24
#     ),
#     naics = "11", # NAICS code for Agriculture, Forestry, Fishing and Hunting
#     .by = c(COUNTY_CODE)
#   ) |>
#   fill(interp_value, .direction = "downup")

# # Sum workers and proprietors to get total ag employment
# ag_total <- ag_interp |>
#   select(
#     year = YEAR,
#     area_id = COUNTY_CODE,
#     COUNTY_NAME,
#     naics,
#     ag_jobs = interp_value
#   )

# write_csv(ag_total, "data/outputs/agriculture_employment.csv")
```

# Combined Series

Next the QCEW data is joined with the NES data to create a combined table of employment and income data for both inc and salary and sole proprietor jobs, with the former coming from the QCEW and the latter from the NES.

```{r}
#| label: create_combined_series

# advance proprietor jobs by one year for most recent missing year based on pct change in wage and salary jobs if there is an equivalent industry - otherwise use the last year available
nes$data <- nes$data %>%
  # filter(year != max(year)) %>%
  bind_rows(
    . |>
      filter(
        naics != "10",
      ) |>
      filter(year == max(year)) |>
      mutate(
        year = year + 1
      ) |>
      left_join(
        qcew_changes,
        by = c("area_id", "naics", "own_code", "year")
      ) |>
      mutate(
        prop = prop * (1 + replace_na(ws_change, 0)),
        inc = inc * (1 + replace_na(ws_change, 0))
      ) %>%
      bind_rows(
        . |> 
          filter(
            lvl == "2"
          ) |> 
          summarize(
          naics = "10",
          lvl = "0",
          prop = sum(prop, na.rm = TRUE),
          inc = sum(inc, na.rm = TRUE),
          .by = c(year, area_id, own_code)
        )
      ) |>
      select(-ws_change)
  )

# create combined series of QCEW and NES data for counties
fcst$tej$series_c <- qcew_data |>
  relocate(lvl, .after = naics) |>
  # add on ag data
  # left_join(
  #   expand_grid(own_code = c("0", "5"), ag_total),
  #   by = c("year", "area_id", "naics", "own_code")
  # ) |>
  # mutate(
  #   ag_jobs = if_else(
  #     naics == "11" & !is.na(ag_jobs),
  #     ag_jobs,
  #     0
  #   )
  # ) |>
  mutate(
    area_id = paste0("08", area_id),
    own_code = str_sub(own_code, 1, 1),
    naics = if_else(naics == "00", "10", naics),
    lvl = if_else(naics == "10", "0", as.character(lvl))
  ) |>
  rename(firmws = total_firms) |>
  filter(
    own_code %in%
      c("1", "2", "3", "5") &
      year >= 2000 &
      lvl %in% c("0", "2", "3")
  ) |>
  full_join(
    nes$data |>
      filter((prop + inc) > 0) |>
      mutate(
        lvl = if_else(naics == "10", "0", lvl)
      ),
    by = c("year", "area_id", "naics", "own_code", "lvl"),
    suffix = c("ws", "p")
  ) |>
  mutate(
    area = if_else(
      area_id == "08000",
      "co",
      paste0("c", str_sub(area_id, -3))
    ),
    area_lvl = if_else(
      area_id == "08000",
      "state",
      "county"
    ),
    across(
      starts_with(c("ws", "inc", "prop", "firmws", "largest_firm")),
      \(x) replace_na(x, 0)
    ),
    jobs = ws + prop,
    #  + ag_jobs,
    inc = incws + incp,
    firms = firmws + prop
  ) |>
  # select(-ag_jobs) |>
  relocate(area, .after = area_id) %>%
  # add consolidate 0 ownership code series
  bind_rows(
    . |>
      # filter((agglvl_code%%10) %in% c(0, 3:5)) |>
      mutate(
        own_code = "0"
      ) |>
      summarize(
        across(
          starts_with(c("jobs", "ws", "inc", "prop", "firmws", "firms")),
          ~ sum(.x, na.rm = FALSE)
        ),
        across(
          starts_with("largest_firm"),
          ~ max(.x, na.rm = TRUE)
        ),
        .by = c(
          "year",
          "area_id",
          "area",
          "area_lvl",
          "naics",
          "own_code",
          "lvl"
        )
      )
  ) |>
  relocate(starts_with("f_"), .after = last_col()) |>
  arrange(year, area_id, naics, own_code)

# create combined series for Regions
fcst$tej$series_r <- fcst$tej$series_c |>
  mutate(county_fips = str_sub(area_id, -3)) |>
  inner_join(
    xwalk_region_tbl |>
      filter(region <= 14) |>
      select(
        county_fips,
        region_code
      )
  ) |>
  mutate(
    area_id = paste0("08", region_code),
    area_lvl = "region",
    area = paste0("r", region_code)
  ) |>
  summarize(
    across(
      starts_with(c("jobs", "ws", "inc", "prop", "firm")),
      ~ sum(.x, na.rm = TRUE)
    ),
    across(
      starts_with("largest_firm"),
      ~ max(.x, na.rm = TRUE)
    ),
    .by = c(
      "year",
      "area_id",
      "area",
      "area_lvl",
      "naics",
      "own_code",
      "lvl"
    )
  ) |>
  arrange(year, area_id, naics, own_code)

# create combined series for Colorado
fcst$tej$series <- bind_rows(
  fcst$tej$series_c,
  fcst$tej$series_r
)

dbWriteTable(
  conn$qcew,
  "total_estimated_jobs",
  fcst$tej$series,
  overwrite = TRUE
)

dbDisconnect(conn$qcew)
dbDisconnect(conn$qcew)

```

# Forecast Preparation

## Total Estimated Jobs

Data next has to be prepared for forecasting in Eviews. This is done at the 2 digit NAICS level, and so data must be output at that level. Government jobs are aggregated into a single industry for each state, local, and federal government, and all other industries are private jobs.

```{r}
#| label: prepare_total_jobs

# load jobs crosswalk from csv
xwalk_jobs <- sdodemog::pull_db_table("xwalk", "forecast_jobs")

fcst$tej$output <- fcst$tej$series |>
  select(
    -starts_with("f_")
  ) |>
  pivot_longer(
    starts_with(c("jobs", "ws", "inc", "prop")),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    naics = if_else(
      own_code %in% c("1", "2", "3", "8") & lvl == "4",
      "100",
      naics
    ),
    .after = naics
  ) |>
  inner_join(
    xwalk_jobs |>
      select(
        naics,
        own_code,
        ind,
        ind_sort
      ) |>
      filter(
        !is.na(ind)
      )
  ) %>%
  bind_rows(
    . |>
      filter(ind %in% c("traderet", "tradewh", "transware", "util")) |>
      mutate(ind = "ttu", naics = "200")
  ) |>
  mutate(
    var = paste(
      variable,
      ind,
      area,
      sep = "_"
    )
  ) |>
  summarize(
    value = sum(value, na.rm = TRUE),
    .by = c(year, area_id, area_lvl, var, naics) #, naics, own_code, lvl, var)
  ) |>
  arrange(year, area_id, naics) |>
  pivot_wider(
    names_from = "year",
    values_from = "value"
  )


# Save the outputx to a CSV file
fn$save_outputs <- function(df, lvl) {
  file_loc <- paste0("data/outputs/", lvl, "_jobs_est.csv")

  df |>
    filter(
      area_lvl == lvl
    ) |>
    select(var, matches("\\d{4}$")) |>
    write_csv(file_loc)

  drive_put(
    file_loc,
    path = as_id(files$gdrive_folder),
    name = basename(file_loc),
    type = "csv"
  )
}

map(c("county", "state", "region"), \(x) fn$save_outputs(fcst$tej$output, x))


```

## Moodys Job Estimates

Moodys Analytics provides national forecasts of various economic variables that are used to guide the Colorado forecasts. While most of these are industry job estimates, other variables are also used that may be useful in forecasting jobs at the local level. These variables are downloaded from the [Moodys Analytics Data Buffet](https://www.economy.com/data) and stored in the DuckDB database for further analysis. The data is downloaded using the `get_moodys_series` function, which retrieves the data for a given year and variable. A crosswalk file is used for mapping the Moodys analytics industry codes to the NAICS codes used in the Colorado estimates.

```{r}
# | label: load_moodys_jobs

# get list of moody's employment variables
get_moodys_token()
check_moodys_token()

# pull list of moody's variables from the crosswalk to pipe into pull
fcst$vars$jobs <- xwalk_jobs |>
  pull(mnemonic)


# add moodys data using sdoecon get_moodys_series function
fcst$moodysjobs$raw <- map(
  #break fcst into chunks of max 25
  expand_grid(
    split(fcst$vars$jobs, seq_along(fcst$vars$jobs) %/% 25),
    as.character(c(1:12))
  ),
  \(x) {
    get_moodys_series(x, vintage = "202508", freq = "204") |>
      convert_moodys()
  }
) |>
  bind_rows()

map(as.character(c(1:12)), \(x) {
  get_moodys_series(
    "FET.IUSA",
    vintage = paste0("2024", x),
    freq = "204"
  ) |>
    convert_moodys() |>
    mutate(
      vintage = paste0("2024", "-", x)
    )
})


sdo_db_write_table(
  fcst$moodysjobs$raw,
  schema = "forecast",
  table = "moodys",
  vintage = "2025-",
  version = "v1",
  overwrite = TRUE,
  latest = TRUE
)

fcst$moodysjobs$data <- pull_db_table(schema = "forecast", table = "moodys") |>
  inner_join(
    xwalk_jobs |>
      filter(
        !is.na(ind)
      ) |>
      mutate(
        var = paste0("jobs_", ind, "_us")
      ) |>
      select(
        mnemonic,
        var
      ),
    by = c("mnemonic")
  ) |>
  relocate(var, .after = mnemonic) |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

files$moodysjobs <- "data/outputs/moodysjob_vars.csv"

# Save the output to a CSV file
fcst$moodysjobs$data |>
  write_csv(files$moodysjobs)


drive_put(
  files$moodysjobs,
  path = as_id(files$gdrive_folder),
  name = basename(files$moodysjobs),
  type = "csv"
)
```

```{r}
# | label: load_moodys_other

xwalk_moodys <- pull_db_table("xwalk", "forecast_moodys")

# pull list of moody's variables from the crosswalk to pipe into pull
fcst$vars$moodys <- xwalk_moodys |>
  pull(mnemonic)


# add moodys data using sdoecon get_moodys_series function
fcst$moodys$raw <- map(
  #break fcst into chunks of max 25
  split(fcst$vars$moodys, seq_along(fcst$vars$moodys) %/% 25),
  \(x) {
    get_moodys_series(x, freq = "204") |>
      convert_moodys()
  }
) |>
  bind_rows()

fcst$moodys$combined <- bind_rows(
  fcst$moodysjobs$raw,
  fcst$moodys$raw
)

sdo_db_write_table(
  fcst$moodys$combined,
  schema = "forecast",
  table = "moodys",
  vintage = "2025-09",
  version = "v1",
  overwrite = TRUE,
  latest = TRUE
)

fcst$moodys$data <- fcst$moodys$raw |>
  inner_join(
    xwalk_moodys |>
      select(
        mnemonic,
        var
      ),
    by = c("mnemonic")
  ) |>
  relocate(var, .after = mnemonic) |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

files$moodys <- "data/outputs/moodys_vars.csv"

# Save the output to a CSV file
fcst$moodys$data |>
  write_csv(files$moodys)


drive_put(
  files$moodys,
  path = as_id(files$gdrive_folder),
  name = basename(files$moodys),
  type = "csv"
)
```

## FRED Variables

A number of additional economic (non direct jobs) variables are used in the forecasting process, and these are downloaded from the [FRED API](https://fred.stlouisfed.org/). These variables are used to provide additional context for the forecasting process, and are stored in the DuckDB database for further analysis. The `get_fred_series` function from the fredr package is used to retrieve the data for a given year and variable. A crosswalk file is used for mapping FRED variables is stored in "data/crosswalks/xwalk_fred.csv".

```{r}
# | label: load_fred_data

# load fred crosswalk from csv
xwalk_fred <- read_csv(
  "data/crosswalks/xwalk_fred.csv"
)

# create laus crosswalk
xwalk_fred_laus <- expand_grid(
  county_fips = unique(xwalk_region_tbl$county_fips),
  fred_id_num = as.character(3:5)
) |>
  left_join(
    tibble(
      fred_id_num = as.character(3:5),
      var = c("ur", "unem", "empl")
    )
  ) |>
  mutate(
    fred_series = paste0(
      "LAUCN08",
      county_fips,
      "000000000",
      fred_id_num,
      "A"
    )
  ) |>
  arrange(county_fips, fred_id_num)

# pull variables from crosswalk to pipe into pull
fcst$fred$raw <- xwalk_fred |>
  pmap(
    \(fred_series, var) {
      fredr(
        fred_series,
        frequency = "a",
        aggregation_method = "avg"
      ) |>
        mutate(
          var = var,
          .before = 1
        )
    },
    .progress = TRUE
  ) |>
  bind_rows()
fcst <- list()

# download county level laus data
# fcst$fred$raw_laus <- xwalk_fred_laus |>
#   filter(
#     county_fips != "014" # exclude Broomfield as it does not have its own series
#   ) |>
#   select(fred_series, var, county_fips) |>
#   pmap(
#     slowly(
#       \(fred_series, var, county_fips) {
#         fredr(
#           fred_series
#         ) |>
#           mutate(
#             county_fips = county_fips,
#             var = var,
#             .before = 1
#           )
#       },
#       rate = rate_delay(0.15)
#     ),
#     .progress = TRUE
#   ) |>
#   bind_rows()

# pull county level laus data from postgres to avoid repeated downloads

fcst$fred$data <- fcst$fred$raw |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 1990
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

# fcst$fred$data_laus <- fcst$fred$raw_laus |>
#   mutate(
#     year = year(date),
#     #modify var to extract text from between the two "_"
#     var = str_extract(var, "(?<=cps_).*(?=_c)")
#   ) |>
#   select(year, county_fips, var, value)

fcst$fred$county_laus <- sdodemog::pull_db_table(
  schema = "econ",
  table = "laus_county"
)

fcst$fred$region_laus <- sdodemog::pull_db_table(
  schema = "econ",
  table = "laus_region"
)


files$fred <- "data/outputs/fred_vars.csv"

# Save the output to a CSV file
fcst$fred$data |>
  write_csv(files$fred)


drive_put(
  files$fred,
  path = as_id(files$gdrive_folder),
  name = basename(files$fred),
  type = "csv"
)
```

## SDO Variables

```{r}
#| label: load_sdo_vars

# get sdo tables from postgres
sdo <- list()

# load pop by sya
sdo$pop_sya <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.county_sya")
) |>
  collect()

sdo$lfpr <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.labor_force_participation")
) |>
  collect()

sdo$profiles <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.county_profiles")
) |>
  collect()

sdo$household_projections <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.household_projections")
) |>
  collect()

sdo$components_change <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM estimates.components_change")
) |>
  collect()


sdo$output <- sdo$pop_sya |>
  filter(countyfips == 0) |>
  mutate(
    age_group = cut(age, c(0, 4, 19, 24, 44, 64, 74, 84, Inf), right = FALSE)
  ) |>
  summarize(
    demo_pop_co = sum(totalpopulation),
    .by = c(year, countyfips, age_group)
  ) |>
  mutate(
    age_group = paste0(
      "pop",
      str_replace_all(as.character(age_group), "\\[|\\)|,|Inf", "")
    )
  ) |>
  pivot_wider(
    names_from = age_group,
    values_from = demo_pop_co
  ) |>
  arrange(year) |>
  left_join(
    sdo$profiles |>
      select(
        countyfips,
        year,
        births,
        deaths,
        netmig = netmigration,
        hh = households,
        hu = totalhousingunits,
        poptot = totalpopulation
      ) |>
      bind_rows(
        sdo$components_change |>
          filter(countyfips == 0 & datatype == "Projection") |>
          select(
            countyfips,
            year,
            births,
            deaths,
            netmig,
            poptot = estimate
          ) |>
          left_join(
            sdo$household_projections |>
              filter(age_group_id == 0 & household_type_id == 0) |>
              select(
                countyfips = area_code,
                year,
                hh = total_households
              ),
            by = c("countyfips", "year")
          )
      )
  ) |>
  left_join(
    sdo$lfpr |>
      rename(year = population_year) |>
      summarize(
        lf = sum(laborforce),
        cni16 = sum(cni_pop_16pl),
        lfpr = lf / cni16,
        .by = c(year)
      )
  ) |>
  arrange(countyfips, year) |>
  pivot_longer(
    cols = -c(year, countyfips),
    names_to = "var",
    values_to = "value"
  ) |>
  mutate(var = paste("demo", var, "co", sep = "_")) |>
  select(-countyfips) |>
  pivot_wider(
    names_from = var,
    values_from = value
  ) |>
  arrange(year)

files$sdo <- "data/outputs/sdo_vars.csv"

sdo$output |>
  write_csv(files$sdo)

drive_put(
  files$sdo,
  path = as_id(files$gdrive_folder),
  name = basename(files$sdo),
  type = "csv"
)

```

## Labor Force Participation Rates by Age

The BLS provides both estimated and projected labor force participation rates by age and sex at the national level, however state and local level data is not available at either the state or local level. In order to estimate labor force participation rates by age and sex for Colorado and counties within it the SDO begins with ACS estimates of labor force participation rates and then applies the BLS projected labor force participation to project forward LFPR.

## LEHD Data

Data from the [LEHD](https://www.census.gov/programs-surveys/lehd.html) program is used to estimate the number of commuters and multiple job holding rates for Colorado. The LEHD data is produced using a separate script, that can be found either on the [Google Drive](https://drive.google.com/drive/folders/1z3nXScGg1KMfK-3bPrmDbOMhmCGlDvmY?usp=drive_link), or on the [process_lehd GitHub repository](https://github.com/gtottenSDO/process_lehd). The Google Drive link is referenced as it already has the necessary files downloaded, which can take some time if not already saved.

Once the LEHD data is downloaded and processed, it is stored both in a duckdb database (found on the Google Drive) and with the final tables saved in Postgres. This program just pulls the final tables from the Postgres database, which can then be loaded into the forecast.

```{r}
#| label: load_lehd

# load LEHD data from postgres into lehd list
lehd <- list()

lehd$commute <- dbReadTable(
  conn$pgsdo,
  Id(
    schema = "econ",
    table = "lehd_net"
  )
)

# load multiple job holding
lehd$mjh <- dbReadTable(
  conn$pgsdo,
  Id(
    schema = "econ",
    table = "lehd_mjh"
  )
)

# combine commute and multiple job holding into single output table
lehd$output <- lehd$commute |>
  summarize(
    county_fips = "co",
    commute = sum(net_commute, na.rm = TRUE),
    .by = c("year")
  ) |>
  left_join(
    lehd$mjh |>
      summarize(
        county_fips = "co",
        primary_jobs = sum(primary_jobs, na.rm = TRUE),
        all_jobs = sum(all_jobs, na.rm = TRUE),
        mjhr = all_jobs / primary_jobs,
        .by = c("year", state_fips)
      ),
    by = c("year", "county_fips")
  ) |>
  filter(state_fips == "08") |>
  arrange(year) |>
  select(
    year,
    county_fips,
    commute,
    mjhr
  ) |>
  pivot_longer(
    cols = -c(year, county_fips)
  ) |>
  mutate(
    var = paste("lehd", name, county_fips, sep = "_")
  ) |>
  select(-county_fips, -name) |>
  pivot_wider(
    names_from = var,
    values_from = value
  )

files$lehd <- "data/outputs/lehd_vars.csv"

lehd$output |>
  write_csv(files$lehd)

drive_put(
  files$lehd,
  path = as_id(files$gdrive_folder),
  name = basename(files$lehd),
  type = "csv"
)
```