---
title: "2024 Jobs Estimates"
format: html
params:
    est_year: 2024
---

# Overview

This document is for producing the `{r} params$est_year` Colorado Jobs Estimates from the Colorado State Demography Office ("SDO").

## Changes

The methodology utilized by the SDO in its 2024 estimates program has been significantly revised for multiple reasons, including:

### BLS Data Availability

Changes in BLS (Bureau of Labor Statistics) data availability require a new primary source of certain employment and income data. In particular, industry specific data is no longer available, and because of this the SDO has decided to instead use the Quarterly Census of Employment and Wages (QCEW) and Current Employment Statistics (CES) data as the primary source of employment and income data. This simplifies the analysis and allows for more consistent data across industries.

Additionally the availability of income data has changed from what has historically been available from the BLS. Additional sources of income data instead must be relied upon. The new methodology is outlined in the Base Industry section below.

### Employment Sources and Industry Classification

In order to simplify the methodology and allow for consistency with other data sources the SDO has decided to use the North American Industry Classification System (NAICS) as the primary industry classification system. This allows for consistency with other data sources and simplifies the analysis. Additionally, the SDO has decided to use the public versions of the QCEW and CES data, rather than relying on the microdata at this time, to again simplify the process. If significant fidelity is lost in the new methodology, the SDO may consider using the microdata in the future.

## Loading Data and Packages

```{r}
library(tidyverse)
library(keyring)
library(DBI)
library(RPostgres)
library(duckdb)
library(dbplyr)
library(censusapi)
library(connections)
library(duckplyr)
library(sdoecon)
library(fredr)
```

Crosswalks are loaded from the SDO Postgres database, using the generic public (non read-write) connection. Connection keys are stored in the `.Renviron` file, which is not included in this repository for security reasons. The Postgres database is accessed using the `DBI` and `RPostgres` packages. Credentials are stored in the `keyring` package, which allows for secure storage of credentials^[See https://keyring.r-lib.org/index.html for more information. Public credentials are stored under the service `"pgsdo"`, for the username `"codemog"`]. The `duckdb` package is used to create a local DuckDB database for processing the data stored as `data/qcew.duckdb`.

```{r}
#| label: connect_to_db

# create connections list
conn <- list()


# connect to SDO Postgres database
conn$pgsdo <- DBI::dbConnect(
  RPostgres::Postgres(),
  dbname = 'dola',
  host = '104.197.26.248',
  port = 5433,
  user = "codemog",
  password = keyring::key_get("pgsdo", "codemog")
)

# Connect to duckdb database of all QCEW data
conn$qcew <- dbConnect(duckdb::duckdb(), "data/qcew.duckdb")

# Optional - use connections for browsing database in compatabile IDE (e.g., RStudio/Positron)
connections::connection_view(conn$pgsdo)
connections::connection_view(conn$qcew)

# create list for functions
fn <- list()

```

```{r}
#| label: load_crosswalks

dbListTables(conn$pgsdo)

# create crosswalks list
crosswalks <- list()

# store crosswalk_region
crosswalks$region <- tbl(
  conn$pgsdo,
  sql("SELECT * FROM econ.crosswalk_region")
)

```

# Total Estimated Jobs

The first step in the process is to estimate the total number of jobs in Colorado for the year `{r params$est_year}`. This is done using the QCEW and CES data, which provides a comprehensive view of employment in the state. Additionally, an estimate of the total number of sole proprietors is derived using the U.S. Census Bureau's Nonemployer Statistics (NES) data, which provides an estimate of the number of sole proprietors in the state.

## QCEW

First QCEW data is downloaded and processed for all years. QCEW data is accessed using the [BLS QCEW CSV Data Slices](https://www.bls.gov/cew/additional-resources/open-data/csv-data-slices.htm) system for accessing CSVs. Data can be downloaded as CSV files using the URL format https://data.bls.gov/cew/data/api/{YEAR}/{QTR}/area/{FIPS5}.csv, where `{YEAR}` is the year of the analysis, starting in 1990 to current, `{QTR}` is `1-4` for a given quarter, or `a` for annual average, and `{FIPS5}` is the 5 digit state plus county FIPS code. A script is provided to download and process the QCEW data for all years, and is available in `r_scripts/process_qcew.R`. The script downloads the data, processes it, and stores it in a DuckDB database `data/qcew.duckdb` for further analysis. Since the process is time consuming, the script is not run in this document, but is available for reference. A Google Drive folder is used for backing up date including both the [raw data](https://drive.google.com/open?id=1xNJTL7PMTY0wDlIUqWh3eE931s62vbh2&usp=drive_fs), and the [processed DuckDB database]().

```{r}
#| label: process_qcew

# source the processing script (uncomment to source)
# source("r_scripts/process_qcew.R")

```

```{r}
#| label: load_qcew

# create list for storing local qcew info
qcew <- list()

# create a table of raw qcew data
qcew$raw <- tbl(conn$qcew, "qcew") |>
  collect()

# create function for creating crosswalk
fn$qcew <- function(qcew_col) {
  code <- paste0(qcew_col, "_code")
  title <- paste0(qcew_col, "_title")

  qcew$raw |>
    select(year, all_of(code), all_of(title)) |>
    distinct() |>
    mutate(year = as.integer(year)) |>
    mutate(yr_fltr = max(year, na.rm = TRUE) == year, .by = all_of(code)) |>
    filter(yr_fltr) |>
    select(-year) |>
    arrange(!!sym(code)) |>
    collect()
}

# apply function to create crosswalks for each QCEW column
crosswalks$qcew <- c("industry", "agglvl", "own") |>
  set_names() |>
  map(fn$qcew)

# create crosswalk for ind and agglvl
crosswalks$qcew$ind_agglvl <- qcew$raw |>
  select(year, starts_with(c("own", "industry", "agglvl"))) |>
  distinct() |>
  mutate(year = as.integer(year)) |>
  mutate(
    yr_fltr = max(year, na.rm = TRUE) == year,
    .by = (c("industry_code"))
  ) |>
  filter(yr_fltr) |>
  select(-year) |>
  collect() |>
  arrange(industry_code, own_code, agglvl_code)

crosswalks$qcew$indlvl <- crosswalks$qcew$ind_agglvl |>
  mutate(
    agglvl_code = if_else(
      industry_code == "10",
      "0",
      agglvl_code
    ),
    indlvl = str_sub(agglvl_code, -1),
  ) |>
  select(industry_code, indlvl) |>
  distinct() |>
  arrange(industry_code) |>
  collect() |>
  arrange(industry_code)


# extract relevant columns and filter agg levels to 50-55 (down to 3 digit NAICS)
qcew$data <- qcew$raw |>
  mutate(
    year = as.integer(year),
    agglvl_code = (agglvl_code)
  ) |>
  filter(
    qtr == "A",
    str_sub(area_fips, 1, 2) == "08", # Colorado
  ) |>
  select(
    year,
    area_fips,
    own_code,
    agglvl_code,
    industry_code,
    f_emp = disclosure_code,
    f_wage = disclosure_code,
    emp = annual_avg_emplvl,
    wage = total_annual_wages
  ) |>
  arrange(year, area_fips, own_code, industry_code)

```

# Sole Proprietors

```{r}
#| label: load_nes_years

# create nes list
nes <- list()

# get available years using censusapi
nes$info <- listCensusApis() |>
  filter(name == "nonemp")

nes$yrs$raw <- nes$info |>
  pull(vintage)

```

Sole proprietor estimates are derived from the U.S. Census Bureau's [Nonemployer Statistics (NES)](https://www.census.gov/programs-surveys/nonemployer-statistics.html) data, which provides an estimate of the number of sole proprietors by county and NAICS. Data is access using the [NES API](https://www.census.gov/data/developers/data-sets/nonemp-api.html). The NES data is released with a lag, and as of the current analysis the most recent available year is `{r} max(nes$yrs$raw)`. The data is retrieved using the [`censusapi`](https://www.hrecht.com/censusapi/) package, which provides a convenient interface to the Census API. The NES data is processed to create a table of sole proprietors by county and NAICS code. The NAICS code is renamed to `industry_code`^[As there are not strict concordances between NAICS years we leave these classified as they were at the time - as with the QCEW. At the industry level there should be minimal changes. The one exception to align with the QCEW is that the QCEW utilizes 2002 NAICS codes for all years prior to . 1997 to 2002 concordances were used, allocating evenly. ] to match the QCEW data, and use the `own_code` of `1` to indicate sole proprietors. The NES data is then stored in the DuckDB database for further analysis.

```{r}
#| label: create_naics_crosswalk

# create crosswalk for NAICS codes
crosswalks$naics <- read_csv_duckdb(
  "data/crosswalks/naics1997_2002.csv",
  options = list(
    all_varchar = "true"
  )
) |>
  collect() |>
  select(
    industry_code = NAICS97,
    NAICS02 = NAICS02
  ) |>
  filter(!is.na(NAICS02)) |>
  add_count(industry_code) |>
  mutate(alloc = 1 / n) |>
  select(-n)

```

```{r}
#| label: load_nes

# load NES variables for most recent year
nes$vars <- listCensusMetadata(
  name = "nonemp",
  vintage = max(nes$yrs$raw),
  type = "variables"
)

fn$get_nes_data <- function(vintage) {
  # get list of variables for vintage
  nes$vars <- listCensusMetadata(
    name = "nonemp",
    vintage = vintage,
    type = "variables"
  ) |>
    filter(
      is.na(predicateOnly) &
        !(name %in% c("STATE", "COUNTY"))
    ) |>
    pull(name)

  # get NES data for a given year
  data <- getCensus(
    name = "nonemp",
    vintage = vintage,
    vars = nes$vars,
    region = "county:*",
    regionin = "state:08"
  ) |>
    mutate(
      vintage = vintage,
      .before = 1
    ) |>
    # rename variables to lowercase
    rename_with(tolower) |>
    # rename naics to match QCEW
    rename_with(
      ~"industry_code",
      matches("naics\\d{2,4}$")
    )

  return(data)
}

nes$raw <- tbl(conn$qcew, "nes") |>
  distinct() |>
  arrange(vintage, state, industry_code, county) |>
  collect()

nes$yrs$db <- nes$raw |>
  pull(vintage) |>
  unique()

nes$yrs$qry <- nes$info |>
  select(vintage) |>
  filter(vintage >= 2000 & !(vintage %in% nes$yrs$db)) |>
  pull(vintage) |>
  unique()

# Store NES data in DuckDB
dbWriteTable(
  conn$qcew,
  "nes",
  map(
    nes$yrs$qry,
    fn$get_nes_data,
    .progress = TRUE
  ) |>
    bind_rows(),
  append = TRUE
)

# Create crosswalk for NES data
crosswalks$nes <- nes$raw |>
  collect() |>
  select(industry_code, sector, subsector, indlevel) |>
  distinct() |>
  arrange(industry_code)


# Create table of relevant NES data
nes$data <- nes$raw |>
  mutate(
    area_fips = paste0(state, county),
    own_code = "5", # treat sole props as private,
    industry_code = if_else(
      industry_code == "00",
      "10",
      industry_code
    )
  ) |>
  left_join(
    crosswalks$qcew$indlvl
  ) |>
  select(
    year = vintage,
    area_fips,
    own_code,
    industry_code,
    indlvl,
    emp = nestab,
    wage = nrcptot,
    f_emp = nestab_f,
    f_wage = nrcptot_f
  ) %>%
  # distinct() |>
  bind_rows(
    . |>
      mutate(
        area_fips = "08000", # Colorado total
      ) |>
      summarize(
        across(
          starts_with(c("emp", "wage")),
          ~ sum(.x, na.rm = FALSE)
        ),
        .by = c("year", "area_fips", "industry_code", "indlvl", "own_code")
      )
  ) |>
  arrange(year, area_fips, industry_code, own_code)


```

# Combined Series

Next the QCEW data is joined with the NES data to create a combined table of employment and income data for both wage and salary and sole proprietor jobs, with the former coming from the QCEW and the latter from the NES.
```{r}
#| label: create_combined_series

# create forecast data list
fcst <- list()

# create combined series of QCEW and NES data
fcst$tej$series <- qcew$data |>
  left_join(
    crosswalks$qcew$indlvl
  ) |>
  relocate(indlvl, .after = industry_code) |>
  select(-agglvl_code) |>
  filter(
    own_code %in%
      c("1", "2", "3", "5") &
      year >= 2000 &
      indlvl %in% as.character(c(0:1, 4:5))
  ) |>
  full_join(
    nes$data |>
      filter((emp + wage) > 0),
    by = c("year", "area_fips", "industry_code", "own_code", "indlvl"),
    suffix = c("ws", "prop")
  ) |>
  mutate(
    across(
      starts_with(c("emp", "wage")),
      \(x) replace_na(x, 0)
    ),
    emp = empws + empprop,
    wage = wagews + wageprop
  ) %>%
  bind_rows(
    . |>
      # filter((agglvl_code%%10) %in% c(0, 3:5)) |>
      mutate(
        own_code = "0"
      ) |>
      summarize(
        across(
          starts_with(c("emp", "wage")),
          ~ sum(.x, na.rm = FALSE)
        ),
        .by = c("year", "area_fips", "industry_code", "own_code", "indlvl")
      )
  ) |>
  relocate(starts_with("f_"), .after = last_col()) |>
  arrange(year, area_fips, industry_code, own_code)


```

# Forecast Preparation

## Total Estimated Jobs
Data next has to be prepared for forecasting in Eviews. This is done at the 2 digit NAICS level, and so data must be output at that level. Government jobs are aggregated into a single industry for each state, local, and federal government, and all other industries are private jobs.

```{r}
#| label: prepare_total_jobs

# load jobs crosswalk from csv
crosswalks$jobs <- read_csv(
  "data/crosswalks/xwalk_jobs.csv",
  col_types = "ccccci"
)


fcst$tej$output <- fcst$tej$series |>
  select(
    -starts_with("f_")
  ) |>
  pivot_longer(
    starts_with(c("emp", "wage")),
    names_to = "variable",
    values_to = "value"
  ) |>
  mutate(
    naics = if_else(
      own_code %in% c("1", "2", "3", "8"),
      "100",
      industry_code
    ),
    .after = industry_code
  ) |>
  inner_join(
    crosswalks$jobs |>
      select(
        naics,
        own_code,
        ind,
        ind_sort
      ) |>
      filter(
        !is.na(ind)
      )
  ) |>
  mutate(
    area = if_else(
      area_fips == "08000",
      "co",
      paste0("c", str_sub(area_fips, -3))
    ),
    var = paste(
      str_replace_all(variable, "(e)mp|(w)age", "\\1\\2"),
      ind,
      area,
      sep = "_"
    )
  ) |>
  summarize(
    value = sum(value, na.rm = TRUE),
    .by = c(year, area_fips, var, ind_sort)
  ) |>
  arrange(year, area_fips, ind_sort) |>
  pivot_wider(
    names_from = "year",
    values_from = "value"
  ) |>
  select(-ind_sort)

# Save the output to a CSV file
fcst$tej$output |>
  filter(
    area_fips == "08000"
  ) |>
  select(-area_fips) |>
  write_csv("data/outputs/co_jobs_est.csv")
```


## Moodys Job Estimates
Moodys Analytics provides national forecasts of various economic variables that are used to guide the Colorado forecasts. While most of these are industry job estimates, other variables are also used that may be useful in forecasting jobs at the local level. These variables are downloaded from the [Moodys Analytics Data Buffet](https://www.economy.com/data) and stored in the DuckDB database for further analysis. The data is downloaded using the `get_moodys_series` function, which retrieves the data for a given year and variable. A crosswalk file is used for mapping the Moodys analytics industry codes to the NAICS codes used in the Colorado estimates. 

```{r}
# | label: load_moodys_jobs

# get list of moody's employment variables
get_moodys_token()
check_moodys_token()

# pull list of moody's variables from the crosswalk to pipe into pull
moodys_vars <- crosswalks$jobs |>
  pull(mnemonic)


# add moodys data using sdoecon get_moodys_series function
fcst$moodys$raw <- map(
  #break fcst into chunks of max 25
  split(moodys_vars, seq_along(moodys_vars) %/% 25),
  \(x) {
    get_moodys_series(x, freq = "204") |>
      convert_moodys()
  }
) |>
  bind_rows()

fcst$moodys$data <- fcst$moodys$raw |>
  left_join(
    crosswalks$jobs |>
      mutate(
        var = paste0("jobs_", ind, "_us")
      ) |>
      select(
        mnemonic,
        var
      ),
    by = c("mnemonic")
  ) |>
  relocate(var, .after = mnemonic) |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 2000
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

# Save the output to a CSV file
fcst$moodys$data |>
  write_csv("data/outputs/moodys_vars.csv")
```

## FRED Variables

A number of additional economic (non direct jobs) variables are used in the forecasting process, and these are downloaded from the [FRED API](https://fred.stlouisfed.org/). These variables are used to provide additional context for the forecasting process, and are stored in the DuckDB database for further analysis. The `get_fred_series` function from the fredr package is used to retrieve the data for a given year and variable. A crosswalk file is used for mapping FRED variables is stored in "data/crosswalks/xwalk_fred.csv".

``` {r}
# | label: load_fred_data

# load fred crosswalk from csv
crosswalks$fred <- read_csv(
  "data/crosswalks/xwalk_fred.csv"
)

# pull variables from crosswalk to pipe into pull
fcst$fred$raw <- crosswalks$fred |>
  pmap(
    \(fred_series, var) {
      fredr(
        fred_series,
        frequency = "a",
        aggregation_method = "avg"
      ) |>
        mutate(
          var = var,
          .after = series_id
        )
    },
    .progress = TRUE
  ) |>
  bind_rows()

fcst$fred$data <- fcst$fred$raw |>
  mutate(
    date = ymd(date)
  ) |>
  filter(
    year(date) >= 2000
  ) |>
  pivot_wider(
    names_from = date,
    values_from = value
  )

# Save the output to a CSV file
fcst$fred$data |>
  write_csv("data/outputs/fred_vars.csv")
```
