### Private Jobs

A separate data frame is created for the next steps in the process using only private ownership. This is for simplicity in creating and tracking data and will be rejoined with the government data at the end of the process.

```{r}
#| label: filter_private_ownership

qcew$data_gov <- qcew$data |>
  filter(own_code != "50")


fn$calc_industry_gaps <- function(df) {
  df |>
    select(
      -any_of(c(
        "ind_emp_reported",
        "ind_emp_calculated",
        "ind_emp_gap",
        "ind_supressed"
      ))
    ) |>
    filter(own_code == "50") |>
    left_join(
      df |>
        filter(own_code == "50" & ind_code == "00") |>
        select(
          year,
          county_fips,
          ind_emp_reported = final_avg_emp
        )
    ) |>
    left_join(
      df |>
        filter(own_code == "50" & !ind_code == "00") |>
        summarize(
          ind_emp_calculated = sum(final_avg_emp, na.rm = TRUE),
          ind_supressed = sum(final_suppress, na.rm = TRUE),
          .by = c(year, county_fips)
        )
    ) |>
    mutate(ind_emp_gap = ind_emp_reported - ind_emp_calculated) |>
    arrange(county_fips, ind_code, year)
}

qcew$data$private <- fn$calc_industry_gaps(qcew$data)


```

#### No job gap

First we check to see if there are any counties where the total private jobs (ind_code == "00") is not suppressed, and the sum of all other industries equals the total. If this is the case then any suppressed values can be filled with 0 jobs. This is done to avoid accidentally filling jobs in later steps when there is no job gap to fill.

```{r}
#| label: suppression_adjust_no_gap

qcew$suppress$no_gap <- qcew$data$private |>
  filter(ind_emp_gap <= 0 & ind_supressed > 0 & final_suppress == 1) |>
  select(year, county_fips, ind_code, own_code) |>
  mutate(
    temp_value = 0,
    temp_flag = 1,
    temp_reason = "No Job Gap - Filled with 0"
  )

#apply overrides
qcew$data$private <- fn$suppression_override(
  qcew$suppress$no_gap,
  qcew$data$private
)

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

fn$check_overrides(qcew$data$private)

```

#### Zero Establishments

Next we replace all suppressed values with 0 establishments to have zero employees, and then if there is only one remaining missing NAICS value for that area we fill that as the difference between the parent total and the sum of the other industries. Columns are added to indicate if suppression was overwritten, the override reason, and the new value. Grouped data frames using group_by are used, rather than within mutate, as the same groupings will be used for the next number of steps.

For now we are only calculating these based on the own_code

```{r}
#| label: suppression_adjust_0_estab

qcew$suppress$no_estabs <- qcew$data$private |>
  filter(own_code == "50" & ind_supressed == "1" & final_suppress == "1") |>
  mutate(
    temp_value = max(ind_emp_gap, 0),
    temp_flag = 1,
    temp_reason = "Only Missing Industry",
    .by = c(year, county_fips, ind_code)
  ) |>
  select(
    year,
    county_fips,
    ind_code,
    own_code,
    final_suppress,
    temp_value,
    temp_flag,
    temp_reason
  )


#apply overrides
qcew$data$private <- fn$suppression_override(
  qcew$suppress$no_estabs,
  qcew$data$private
)

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# check overrides
fn$check_overrides(qcew$data$private)

```

### Interpolation

Next we do a simple linear interpolation for any remaining suppressed values, using the imputeTS::na_interpolation function with a max_gap of 2 (so this does not fill overly long gaps).

```{r}
#| label: simple_interpolation

qcew$suppress$simple_interp <- qcew$data$private |>
  arrange(county_fips, ind_code, own_code, year) |>
  mutate(
    na_count = sum(is.na(final_avg_emp)),
    n_years = n(),
    all_na = na_count >= n_years - 1,
    .by = c(county_fips, ind_code, own_code)
  ) |>
  filter(!all_na) |>
  mutate(
    init_value = if_else(
      final_suppress == 1,
      na_interpolation(final_avg_emp, option = "linear", maxgap = 5),
      NA
    ),
    temp_flag = if_else(!is.na(init_value), 1, NA_real_),
    temp_reason = if_else(
      !is.na(init_value),
      "Simple Linear Interpolation",
      NA_character_
    ),
    .by = c(county_fips, ind_code)
  ) |>
  # adjust for county totals
  mutate(
    init_total = sum(init_value, na.rm = TRUE),
    init_share = if_else(
      init_total > 0,
      init_value / init_total,
      0
    ),
    init_count = sum(temp_flag == 1, na.rm = TRUE),
    init_suppress = sum(final_suppress),
    init_allocate_all = init_count == init_suppress,
    .by = c(year, county_fips)
  ) |>
  #adjust temp_value to not be larger than the employment gap
  mutate(
    temp_value = round(
      if_else(
        !is.na(init_value) & (init_allocate_all | init_total >= ind_emp_gap),
        ind_emp_gap * init_share,
        init_value
      ),
      0
    ),
  ) |>
  select(
    year,
    county_fips,
    countyname,
    own_code,
    ind_code,
    ind_title,
    estab,
    final_avg_emp,
    ind_emp_reported,
    ind_emp_calculated,
    ind_emp_gap,
    final_suppress,
    temp_value,
    temp_flag,
    temp_reason,
    starts_with("init_")
  ) |>
  arrange(
    county_fips,
    ind_code,
    year
  ) |>
  filter(temp_flag == 1) |>
  mutate(
    final_total = sum(temp_value, na.rm = TRUE),
    .by = c(year, county_fips),
    .after = ind_emp_gap
  )


#apply overrides
qcew$data$private <- fn$suppression_override(
  qcew$suppress$simple_interp,
  qcew$data$private
)

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# fill 1 remaining suppressed gaps
qcew$data$private <- qcew$data$private |>
  mutate(
    final_avg_emp = if_else(
      final_suppress == 1 & ind_supressed == 1,
      ind_emp_gap,
      final_avg_emp
    ),
    override_flag = if_else(
      final_suppress == 1 & ind_supressed == 1,
      1,
      override_flag
    ),
    override_reason = if_else(
      final_suppress == 1 & ind_supressed == 1,
      "Fill Single Industry - Interpolation",
      override_reason
    ),
    final_suppress = if_else(
      final_suppress == 1 & ind_supressed == 1,
      0,
      final_suppress
    ),
  )

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# check overrides
fn$check_overrides(qcew$data$private)

```

### Region Shares

Next we calculate the share of employment by industry within the region, and allocate the remaining job gaps based on those shares.

```{r}
#| label: region_shares

# calculate region shares
qcew$suppress$region_shares <- qcew$data$private |>
  filter(ind_supressed == 0) |>
  summarize(
    ind_emp_region = sum(final_avg_emp, na.rm = FALSE),
    total_emp_region = sum(ind_emp_reported, na.rm = FALSE),
    ind_share_region = ind_emp_region / total_emp_region,
    .by = c(year, ind_code, region_code)
  ) |>
  arrange(year, region_code, ind_code) |>
  mutate(
    total_industries = n(),
    .by = c(year, region_code)
  )

# check that region shares sum to 1
qcew$suppress$region_shares |>
  filter(ind_code != "00") |>
  summarize(
    total = sum(ind_share_region, na.rm = TRUE),
    .by = c(year, region_code)
  ) |>
  filter(abs(total - 1) > 0.01)

# allocate remaining gaps based on region shares
qcew$suppress$region_allocate <- qcew$data$private |>
  filter(final_suppress == 1) |>
  left_join(
    qcew$suppress$region_shares,
    by = c("year", "ind_code", "region_code")
  ) |>
  mutate(
    temp_value = round(ind_emp_gap * ind_share_region, 0),
    temp_flag = if_else(!is.na(temp_value), 1, NA_real_),
    temp_reason = if_else(
      !is.na(temp_value),
      "Region Share Allocation",
      NA_character_
    ),
    .by = c(year, county_fips)
  ) |>
  select(
    year,
    county_fips,
    ind_code,
    own_code,
    final_suppress,
    temp_value,
    temp_flag,
    temp_reason
  ) |>
  filter(!is.na(temp_value))

#apply overrides
qcew$data$private <- fn$suppression_override(
  qcew$suppress$region_allocate,
  qcew$data$private
)

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# fill 1 remaining suppressed gaps
qcew$data$private <- qcew$data$private |>
  mutate(
    final_avg_emp = if_else(
      final_suppress == 1 & ind_supressed == 1,
      ind_emp_gap,
      final_avg_emp
    ),
    override_flag = if_else(
      final_suppress == 1 & ind_supressed == 1,
      1,
      override_flag
    ),
    override_reason = if_else(
      final_suppress == 1 & ind_supressed == 1,
      "Fill Single Industry - Region Share",
      override_reason
    ),
    final_suppress = if_else(
      final_suppress == 1 & ind_supressed == 1,
      0,
      final_suppress
    ),
  )

# update industry gaps
qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# check overrides
fn$check_overrides(qcew$data$private)


```

Finally this is repeated with state shares for the remaining gaps.

```{r}
#| label: state_shares

# calculate state shares
qcew$suppress$state_shares <- qcew$data$private |>
  filter(county_fips == "000") |>
  summarize(
    ind_emp_state = sum(final_avg_emp, na.rm = FALSE),
    total_emp_state = sum(ind_emp_reported, na.rm = FALSE),
    ind_share_state = ind_emp_state / total_emp_state,
    .by = c(year, ind_code)
  ) |>
  arrange(year, ind_code) |>
  mutate(
    total_industries = n(),
    .by = c(year)
  )

# check that state shares sum to 1
qcew$suppress$state_shares |>
  filter(ind_code != "00") |>
  summarize(total = sum(ind_share_state, na.rm = TRUE), .by = c(year)) |>
  filter(abs(total - 1) > 0.01)

# allocate remaining gaps based on state shares
qcew$suppress$state_allocate <- qcew$data$private |>
  filter(final_suppress == 1) |>
  left_join(
    qcew$suppress$state_shares,
    by = c("year", "ind_code")
  ) |>
  mutate(
    temp_value = round(ind_emp_gap * ind_share_state, 0),
    temp_flag = if_else(!is.na(temp_value), 1, NA_real_),
    temp_reason = if_else(
      !is.na(temp_value),
      "State Share Allocation",
      NA_character_
    ),
    .by = c(year, county_fips)
  ) |>
  select(
    year,
    county_fips,
    ind_code,
    own_code,
    final_suppress,
    temp_value,
    temp_flag,
    temp_reason
  ) |>
  filter(!is.na(temp_value))

#apply overrides
qcew$data$private <- fn$suppression_override(
  qcew$suppress$state_allocate,
  qcew$data$private
)

# update industry gaps

qcew$data$private <- fn$calc_industry_gaps(qcew$data$private)

# check overrides

fn$check_overrides(qcew$data$private)

```