---
title: "Suppressing Jobs Data"
format: html
---

Jobs data estimates needs to be supressed according to BLS guidelines according to the 80/3 rule. This rule requires that that if a single firm is 80% or greater of the total employment in an industry within a given area, or if there are 3 or fewer firms in an industry within a given area, then the data must be suppressed.

# Setup

First we load the necessary libraries and data.

```{r}
library(tidyverse)
library(duckdb)
library(dbplyr)

```

# Connect to DuckDB Database

Load the database with the jobs data from the `data/qcew.duckdb` file created in the initial estimates calculations.

```{r}
#| label: load-db

con <- dbConnect(duckdb::duckdb(), dbdir = "data/qcew.duckdb")
connections::connection_open(duckdb::duckdb(), "data/qcew.duckdb")

estimates_raw <- tbl(con, "total_estimated_jobs")
```

# Apply Suppression Rules

Now we apply the suppression rules to the data.

```{r}
#| label: apply-suppression

suppressed_estimates <- estimates_raw %>%
  group_by(year, quarter, area_fips, industry_code) %>%
  mutate(
    total_employment = sum(estimated_jobs, na.rm = TRUE),
    firm_count = n(),
    max_firm_employment = max(estimated_jobs, na.rm = TRUE),
    percent_max_firm = (max_firm_employment / total_employment) * 100,
    suppress = if_else(percent_max_firm >= 80 | firm_count <= 3, TRUE, FALSE),
    estimated_jobs_suppressed = if_else(suppress, NA_real_, estimated_jobs)
  ) %>%
  select(-total_employment, -firm_count, -max_firm_employment, -percent_max_firm, -suppress)

```