---
title: "Suppressing Jobs Data"
format: html
---

Jobs data estimates needs to be supressed according to BLS guidelines according to the 80/3 rule. This rule requires that that if a single firm is 80% or greater of the total employment in an industry within a given area, or if there are 3 or fewer firms in an industry within a given area, then the data must be suppressed.

# Setup

First we load the necessary libraries and data.

```{r}
library(tidyverse)
library(duckdb)
library(dbplyr)
library(DBI)
library(dbplyr)
library(sdodemog)

```

# Connect to DuckDB Database

Load the database with the jobs data from the `data/qcew.duckdb` file created in the initial estimates calculations.

```{r}
#| label: load-db

con <- dbConnect(duckdb::duckdb(), dbdir = "data/qcew.duckdb")
connections::connection_open(duckdb::duckdb(), "data/qcew.duckdb")

estimates_raw <- tbl(con, "total_estimated_jobs")

qcew <- tbl(con, "qcew")
```


# Apply Suppression Rules

Now we apply the suppression rules to the data.

```{r}
#| label: apply-suppression

suppressed_estimates <- estimates_raw %>%
  filter(
    (own_code %in% c("0", "1", "2", "3") & ind_lvl == "0") |
      own_code == 5 & ind_lvl == "2"
  ) |>
  left_join(
    qcew |>
      select(
        year,
        area_fips,
        own_code,
        ind_code = industry_code,
        annual_avg_emplvl
      )
  ) |>
  mutate(
    ind_code = if_else(
      own_code %in% c("1", "2", "3"),
      own_code,
      ind_code
    ),
    ind_lvl = if_else(
      own_code %in% c("1", "2", "3"),
      "2",
      ind_lvl
    )
  ) |>
  collect() |>
  select(-contains("ws"), -contains("inc"), -starts_with("f_")) |>
  relocate(largest_firm, .after = firms) |>
  mutate(
    largest_firm_share = largest_firm / jobs,
    qcew_diff = abs(jobs - annual_avg_emplvl),
    qcew_override = qcew_diff > 1,
    initial_suppression = if_else(
      (largest_firm_share >= 0.8 | firms <= 3) &
        qcew_override,
      TRUE,
      FALSE
    )
  ) |>
  mutate(
    suppression_count = sum(initial_suppression),
    min_ind = ind_code[which.min(replace_na(jobs, Inf))],
    .by = c(area_fips, ind_lvl, year)
  )

# find
# find the industry code for the industry with lowest number of jobs of all industries including those that are not suppressed
suppress_additional <- suppressed_estimates |>
  filter(suppression_count == 1 & !initial_suppression) |>
  mutate(
    min_ind = ind_code[which.min(replace_na(jobs, Inf))],
    .by = c(area_fips, ind_lvl, year)
  ) |>
  filter(ind_code == min_ind)

final_suppressed_estimates <- suppressed_estimates |>
  left_join(
    suppress_additional |>
      select(area_fips, ind_lvl, year, ind_code) |>
      mutate(additional_suppression = TRUE),
    by = c("area_fips", "ind_lvl", "year", "ind_code")
  ) |>
  mutate(
    additional_suppression = replace_na(additional_suppression, FALSE),
    final_suppression = initial_suppression | additional_suppression,
    final_jobs = if_else(final_suppression, NA_real_, jobs),
    largest_firm = if_else(final_suppression, NA_real_, largest_firm)
  )

suppressed_totals <- final_suppressed_estimates |>
  filter(final_suppression) |>
  summarize(
    total_jobs = sum(jobs, na.rm = TRUE),
    .by = c(year, area_lvl, area_fips, ind_lvl)
  ) |>
  arrange(year, area_lvl, area_fips, ind_lvl)

supressed_totals_by_year_lvl <- suppressed_totals |>
  summarize(
    total_jobs = sum(total_jobs, na.rm = TRUE),
    .by = c(year, area_lvl, ind_lvl)
  ) |>
  arrange(year, area_lvl, ind_lvl)

final_jobs_table <- final_suppressed_estimates |>
  filter(ind_lvl %in% c("0", "2") & year >= 2001) |>
  mutate(area_id = str_sub(area_fips, -3)) |>
  select(
    year,
    area_id,
    ind_code,
    jobs = final_jobs
  )
```

``` {r}
# add to database
sdo_db_write_table(
  final_jobs_table,
  schema = "econ",
  table = "jobs_estimates",
  vintage = "2024",
  version = "v4",
  latest = TRUE,
  overwrite = TRUE
)

dbDisconnect(con)
```

# Compare to prior Estimates

Now we can compare the suppressed estimates to the prior estimates to see the impact of the suppression.

```{r}
#| label: compare-estimates

```
